---
title: "The Fox News Effect"
output:
  pdf_document: default
  html_document: default
date: "5 August 2015"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(error=FALSE,comment='')
```
 

Recently, many scholars have been interested in quantifying the effect
of the national news media on the behavior of electorate.
Understanding how the national media environment affects the election
results is critically important, but also notoriously difficult. In
particular, even if an association between media coverage and election
outcomes exists, it is difficult to identify whether this association
is due to the media's influence over the voters or the result of the
media adjusting its contents to the preferences of voters.

In this exercise, we will consider the entry of the Fox News Channel
(hereafter Fox News) into the television market in the late
1990s. This exercise is based on the following study: Stefano
  DellaVigna and Ethan Kaplan (2007). ["The Fox News Effect: Media
  Bias and Voting."](https://doi.org/10.1162/qjec.122.3.1187) *Quarterly Journal of Economics*,
  122:3, pp.1187-1234.  
  
Note that due to the nature of
negotiations between cable companies and television networks, adding a
new channel to the line-up of a cable company may take a long time.
For this reason, in contrast to what many people might expect, the Fox
News was not able to enter conservative media markets first.  We will
be looking at some of the differences between the towns that initially
did not receive Fox News, and those that did.  The data set is in the
csv file `foxnews.csv`. It contains information for 10,126 towns
across 28 states in the United States:

--------------------------------------------------------------------------------
 Name                 Description
 -------------------- ---------------------------------------------------------
 `town`               Town name
 
 `state`              State in which the town is located
 
 `subrf2000`          Share of Fox News subscribers in 2000
 
 `gopvoteshare2000`   Two-party vote share for the Republicans (2000 Presidential 
                      election)
 
 `gopvoteshare1996`   Two-party vote share for the Republicans (1996 Presidential 
                      election)
 
 `gopvoteshare1992`   Two-party vote share for the Republicans (1992 Presidential 
                      election)
 
 `college1990`        Proportion of population with a college degree in 1990 
 
 `male1990`           Proportion of male population in 1990
 
 `black1990`          Proportion of black population in 1990
 
 `hisp1990`           Proportion of hispanic population in 1990
 
 `income1990`         Median income in 1990
 
 `logincome1990`      Median income in 1990 on the logarithmic scale 
--------------------------------------------------------------------------------


## Question 1

We will investigate whether there are any systematic differences
  in the distribution of some key pre-treatment variables
  (`hs1990`, `black1990`, `hisp1990`, `male1990`, `logincome1990`)
  between towns that received Fox News as compared to those that did
  not. First, create a new variable called `foxnews2000` that
  takes the value of 1 if the share of Fox News subscribers in a given
  town is strictly larger than 0 and equals 0 otherwise.  Create five
  Quantile-Quantile plots to assess the similarity of the
  distributions for towns with and without subscribers across the five
  variables (`hs1990`, `black1990`, `hisp1990`, `male1990`,
    `logincome1990`). Interpret the results. Are there any consistent
  patterns of differences between the two groups of towns? What do the
  plots tell you about our ability to make causal inferences regarding
  the effect of Fox News on the election outcome?

## Answer 1

```{r}
par(cex = 1.5)
## Read the data
foxnews <- read.csv("data/foxnews.csv")

## Fox News Indicator
foxnews$foxnews2000 <- ifelse(foxnews$subrf > 0, 1, 0)

## Proportion of College graduates in 1990
qqplot(foxnews$college1990[foxnews$foxnews2000 == 1], 
       foxnews$college1990[foxnews$foxnews2000 == 0],
       xlab = "Towns with access to Fox News",
       ylab = "Towns with no access to Fox News",
       main = "Proportion of College graduates in 1990") 
abline(0,1, col = "red")

## Proportion of Blacks in 1990 
qqplot(foxnews$black1990[foxnews$foxnews2000 == 1], 
       foxnews$black1990[foxnews$foxnews2000 == 0],
       xlab = "Towns with access to Fox News",
       ylab = "Towns with no access to Fox News",
       main = "Proportion of Blacks in 1990") 
abline(0,1, col = "red")

## Proportion of Hispanics in 1990 
qqplot(foxnews$hisp1990[foxnews$foxnews2000 == 1],
       foxnews$hisp1990[foxnews$foxnews2000 == 0],
       xlab = "Towns with access to Fox News",
       ylab = "Towns with no access to Fox News",
       main = "Proportion of Hispanics in 1990") 
abline(0,1, col = "red")

## Proportion of Males in 1990
qqplot(foxnews$male1990[foxnews$foxnews2000 == 1], 
foxnews$male1990[foxnews$foxnews2000 == 0],
       xlab = "Towns with access to Fox News",
       ylab = "Towns with no access to Fox News",
       main = "Proportion of Males in 1990") 
abline(0,1, col = "red")

## Logarithm of the Median Income in 1990
qqplot(foxnews$logincome1990[foxnews$foxnews2000 == 1], 
       foxnews$logincome1990[foxnews$foxnews2000 == 0], 
       xlab = "Towns with access to Fox News",
       ylab = "Towns with no access to Fox News",
       main = "Logarithm of the Median Income in 1990") 
abline(0,1, col = "red")
```

The distributions for the proportions of high-school graduates
and proportion of males are nearly identical across the towns that
received the Fox News signal as compared to those that did not. The 
identity is not perfect: for example, the lower quantiles for the
proportion of high school graduates are more skewed toward higher values
for towns with no access to Fox News, while the reverse is true for the
upper quantiles. There is more imbalance along the other three covariates.
For the logarithm of the median income, the quantiles lie consistently
below the 45-degree line, indicating that the income distribution for 
towns that received the Fox News signal is systematically shifted to 
higher values as compared to those that did not. Similar, though less
pronounced, imbalance also obtains for the proportions of Hispanic and 
Black residents. These differences between the distributions confirm
that exposure to Fox News cannot be considered entirely random, but 
the feasibility of the as-if random assumption can be defended due to the 
small magnitude of the imbalance on the observed covariates. These
differences imply that any estimate of the causal effect of exposure to
Fox News should, as a further check on our analysis, account for any differences in 
pre-treatment values on the outcome variable across towns that received
the Fox News signal and those that did not. The difference-in-differences
design would be an appropriate choice for that, subject to the validity
of the parallel trend assumption.

## Question 2

We further examine whether there are any clear differences
  between those towns that did receive Fox News and those that did
  not.  To do this, apply the $k$-means algorithm with two clusters to
  the five variables you analyzed in the previous question.  Be sure
  to remove any missing values and scale each variable so that their
  means are zero and standard deviations are one, before applying the
  algorithm.  What is the distribution of the clusters with respect to
  whether or not towns received Fox News?  What are the
  characteristics of each cluster?  Explain how this analysis answers
  the question about our ability to make causal inferences about the
  electoral effect of Fox News.

## Answer 2

```{r}
par(cex = 1.5)
## Setting a seed for replicability
set.seed(12345)
## Subset the data
foxnews.subset <- na.omit(foxnews[ ,c("foxnews2000", "male1990",
                                      "hisp1990", "logincome1990",
                                      "black1990","hisp1990")])
## Scale the data
foxnews.subset.scaled <- scale(foxnews.subset)
## k-means
cluster.2 <- kmeans(foxnews.subset.scaled[ ,2:6], centers = 2)
## Means for each variable by cluster
cluster.2$centers
## Means on the original scale
colMeans(foxnews.subset[cluster.2$cluster == 1, 2:6])
colMeans(foxnews.subset[cluster.2$cluster == 2, 2:6])
## Cross Table Fox vs Clusters
fox.clusters <- table(cluster.2$cluster, 
                      foxnews.subset$foxnews2000) 
fox.clusters
## The rate of exposure to Fox News is similar across clusters
fox.clusters[1,2]/sum(fox.clusters[1,])
fox.clusters[2,2]/sum(fox.clusters[2,])
```

It can be noted that if compared to cluster 2, towns in cluster 1 have
around the same number of males, less hispanics and blacks, while
their income is larger. However, in terms of the presence of Fox News
across these two clusters, we can see it is quite similar in relative
terms i.e., around 18 percent of the towns in each cluster received
the Fox News signal, while the rest did not.  

## Question 3

We begin to examine the relationship between 
  the exposure to Fox News in 2000 and the change in the GOP's vote share 
  from the 1996 to the 2000 Presidential election. First, create a new 
  variable that measures the difference between the Republican vote share 
  in 2000 and in 1996.  Compute the correlation between this new variable 
  and `subrf2000` and provide an interpretation of the result. 
  
## Answer 3

```{r}
par(cex = 1.5)
## Difference in the vote shares 2000 - 1996
foxnews$voteshare.diff <- foxnews$gopvoteshare2000 - 
                          foxnews$gopvoteshare1996

## Correlation:
corr <- cor(foxnews$voteshare.diff, foxnews$subrf2000, 
            use = "complete.obs")
corr
```

The correlation between the two variables stands at `r round(corr,
  digits=3)`, which suggests a weak negative association. This means
that greater exposure to Fox News corresponds to (but does not
necessarily bring about!) a lower electoral gain for the
Republicans. The scatterplot helps to visualize this conclusion.

## Question 4

We now estimate the causal effect of Fox News on the
  Republicans' vote share. For this question, use `foxnews2000`
  to measure exposure to Fox News. Interpret the results. What
  estimation strategy did you use to identify this causal effect? What
  is the assumption required for this analysis to be valid?  Interpret
  this assumption in the context of this particular question. In your
  view, how credible is this assumption?  Use the 1992 and 1996
  election outcomes, both of which took place before the creation of the Fox News
  channel, to probe the credibility of the assumption.

## Answer 4

```{r}
par(cex = 1.5)

## Difference-in-Differences
DiD.treat <- mean(foxnews$voteshare.diff[foxnews$foxnews2000 == 1], 
                  na.rm = TRUE) 
DiD.control <- mean(foxnews$voteshare.diff[foxnews$foxnews2000 == 0], 
                    na.rm = TRUE)

DiD <- DiD.treat - DiD.control
DiD

## Placebo Test
foxnews$voteshare.diff.2 <-  foxnews$gopvoteshare1996 - 
                                foxnews$gopvoteshare1992

Placebo.t <- mean(foxnews$voteshare.diff.2[foxnews$foxnews2000 == 1], 
                  na.rm = TRUE) 
Placebo.c <- mean(foxnews$voteshare.diff.2[foxnews$foxnews2000 == 0], 
                    na.rm = TRUE)

Placebo <- Placebo.t - Placebo.c
Placebo
```

According to our difference-in-differences estimate 
the entry of Fox News made things worse for the Republicans
as it reduced the vote share in favor of the GOP by 
more than 1 percentage point. 

The main assumption behind difference-in-differences is
the 'parallel paths' assumption. Basically, what we
need to assume is that in the absence of the treatment, 
any pre-existing differences in the outcome of interest
would have remained constant across treatment and
control groups. The latter is key to identify the
effect of Fox News, as any movement beyond
the pre-existing differences in towns that did
and did not received the signal can be attributed to 
the entry of Fox News. Given that we could not find
any major difference in other relevant covariates
between towns that received the Fox News signal and 
towns that did not, it seems reasonable to make this
assumption. 

To assess the validity of this assumption, we perform
a 'placebo test.' We do so to discard the possibility
that the effect we found is not due to an unobserved
trend in the electorate agaings the Republicans. In other 
words, Fox News should not have an impact before its
entry to the cable market. If we do find an effect,
then there might be something different in the towns we
are comparing. As it can be noted above, the placebo
test reveals no effect on difference in the vote shares for the
Republicans from 1992 to 1996, when comparing towns
that received the signal of Fox News in the year 2000 with
towns that did not.

## Question 5

We further divide the towns that received Fox News into three
  groups based on the share of Fox News subscribers.  Among the towns
  who received Fox News, the 'High exposure' group represents the
  group of towns whose share of subscribers is greater than or equal
  to the 66 percentile (among those who received Fox News).  In
  contrast, the 'Low exposure' group represents the group of towns
  whose share of subscribers is less than or equal to the 33
  percentile (among those who received Fox News).  Conduct the same
  analysis as in the previous question but separately for the 'High
  exposure' and 'Low exposure' groups where the control group is
  the 'No exposure' group.  Interpret the results.
  
## Answer 5

```{r}
par(cex = 1.5)

## 33rd Percentile
p33 <- quantile( foxnews$subrf[ foxnews$subrf > 0 ], 
                 probs = c(0.33), na.rm = T )

p66 <- quantile( foxnews$subrf[ foxnews$subrf > 0 ], 
                 probs = c(0.66), na.rm = T )

## Fox News by level of exposure
foxnews$foxnews2000.v2 <- ifelse(foxnews$subrf >= p66, 2, 
                                ifelse( foxnews$subrf > 0 
                                        & foxnews$subrf <= p33,
                                        1, 0))
## Table of counts
table(foxnews$foxnews2000.v2)

## Difference in Differences 
diff.t <- mean(foxnews$voteshare.diff[foxnews$foxnews2000.v2 == 2], 
                       na.rm = T) 
diff.c <-mean(foxnews$voteshare.diff[foxnews$foxnews2000.v2 == 0], 
                       na.rm = T)
did.high.nofox <- diff.t - diff.c
did.high.nofox

diff.t2 <- mean(foxnews$voteshare.diff[foxnews$foxnews2000.v2 == 1], 
                       na.rm = T )
diff.c2 <- mean(foxnews$voteshare.diff[foxnews$foxnews2000.v2 == 0], 
                       na.rm = T )
did.low.nofox <- diff.t2 - diff.c2
did.low.nofox
```

The diff-in-diff estimate for the 'Low exposure' treatment is only
`r round(did.low.nofox, digits=5)`, which is considerably
lower, in absolute terms, than the estimate calculated in the previous
question. However, the estimate for the 'High exposure' treatment is
slightly greater in absolute terms, measuring at
`r round(did.high.nofox, digits=5)`. Notice that the direction
of the effect is consistent across the three estimates. This serves as
a further check on the validity of our analysis, confirming the
earlier conclusion that exposure to Fox News marginally reduced the
Republicans' share of the vote in the 2000 Presidential election, as
compared with the 1996 election.

## Question 6

Finally, we consider the effect of having access to Fox News (as
  measured by `foxnews2000`) on the Republican vote share for
  each state.  Repeat the analysis you have done in Question 4 for
  each state and compute a state-specific estimate of the Fox News
  effect (whenever possible). Create a histogram of state-specific
  effects to examine how much the magnitude of the Fox News effect
  varies across states.  Interpret the results.  Finally, compare the
  average effect across states with the estimate you obtained in
  Question 4. What does this comparison suggest about the validity of
  the assumption made in Question 4?
  
## Answer 6

```{r}
## Mean for Towns that received Fox News 
fox.st <- 
    tapply(foxnews$voteshare.diff[foxnews$foxnews2000 == 1], 
           foxnews$state[ foxnews$foxnews2000 == 1], mean, 
           na.rm = TRUE)

## Mean for Towns that did not received Fox News 
nofox.st <- tapply(foxnews$voteshare.diff[foxnews$foxnews2000 == 0], 
                   foxnews$state[foxnews$foxnews2000 == 0], 
                   mean, na.rm = TRUE)

diff.by.state <- fox.st - nofox.st
diff.by.state
hist(diff.by.state, freq = FALSE, 
     main = "Distribution of the Diff-in-Diffs estimates \n 
            across States",
     xlab = "Diff-in-Diffs estimates across states")

comparison <- rbind(DiD,mean(diff.by.state, na.rm = TRUE))
rownames(comparison) <- c("DiD from Question 4", 
                          "Average DiD across States")

comparison

## We can also use a 'for loop' to solve this 
## question though tapply() is better

state.names <- unique(foxnews$state)
diff.by.state.v2 <- rep(NA, length(state.names))
length(state.names)

for(i in 1:length(state.names)) {
    control.group <- 
        mean(foxnews$voteshare.diff[foxnews$foxnews2000 == 0 & 
             foxnews$state == state.names[i]], 
             na.rm = TRUE)
    
    treat.group <- 
        mean(foxnews$voteshare.diff[foxnews$foxnews2000 == 1 & 
             foxnews$state == state.names[i]], 
             na.rm = TRUE)
    diff.by.state.v2[i] <- treat.group - control.group  
}
names(diff.by.state.v2) <- state.names
diff.by.state.v2

mean(diff.by.state.v2, na.rm = TRUE)
```

As it can be noted above, the effect was not homogenous across states.
There were some states that experienced and increase in the
vote share in favor of the Republicans e.g., Alabama, Hawaii, 
North Dakota, Vermont;
while for some other states like California, New York, and 
Wyoming, the effect was really small.
The rest of the states experienced a decrease in
their support for the Republicans. Note that in the case of
New Hampshire, 
there are only towns in the control group, which makes it 
not possible to estimate the
effect of Fox News on the Republicans' vote share. 
That is why the difference-in-differences estimate from Question 4 is slightly different 
from the average difference-in-differences estimate across states - in Question 4
that information was used, while in this
question is not possible to use it.
