---
title: "Understanding World Population Dynamics"
output:
  pdf_document: default
  html_document: default
date: "5 August 2015"
---

Understanding population dynamics is important for many areas of
social science.  We will calculate some basic demographic quantities
of births and deaths for the world's population from two time periods:
1950 to 1955 and 2005 to 2010.  We will analyze the following CSV data
files - `Kenya.csv`, `Sweden.csv`, and `World.csv`.  Each file
contains population data for Kenya, Sweden, and the world,
respectively.  The table below presents the names and descriptions of
the variables in each data set.

--------------------------------------------------------------------------------
 Name                 Description
 -------------------- ----------------------------------------------------------
 `country`            Abbreviated country name
 
 `period`             Period during which data are collected
 
 `age`                Age group 
 
 `births`             Number of births in thousands (i.e., number of children 
                      born to women of the age group) 
 
 `deaths`             Number of deaths in thousands 
 
 `py.men`             Person-years for men in thousands 
 
 `py.women`           Person-years for women in thousands
--------------------------------------------------------------------------------

Source: United Nations, Department of Economic and Social Affairs, Population Division (2013). *World Population Prospects: The 2012 Revision, DVD Edition.*

The data are collected for a period of 5 years where *person-year* is
a measure of the time contribution of each person during the
period. For example, a person that lives through the entire 5 year
period contributes 5 person-years whereas someone who only lives
through the first half of the period contributes 2.5 person-years.
Before you begin this exercise, it would be a good idea to directly
inspect each data set.  In R, this can be done with the `View`
function, which takes as its argument the name of a `data.frame` to be
examined.  Alternatively, in RStudio, double-clicking a `data.frame`
in the `Environment` tab will enable you to view the data in a
spreadsheet-like view.

```{r}
## load the data set
Sweden <- read.csv("data/Sweden.csv")
Kenya <- read.csv("data/Kenya.csv")
World <- read.csv("data/World.csv")
```

## Question 1

We begin by computing *crude birth rate* (CBR) for a given
  period.  The CBR is defined as:
  $$ 
    \text{CBR} 
     =  \frac{\text{number of births}}{\text{number of person-years lived}}
  $$
  
  Compute the CBR for each period, separately for Kenya, Sweden, and
  the world.  Start by computing the total person-years, recorded as a
  new variable within each existing `data.frame` via the
  `$` operator, by summing the person-years for men and women.
  Then, store the results as a vector of length 2 (CBRs for two
  periods) for each region with appropriate labels.  You may wish to
  create your own function for the purpose of efficient programming.
  Briefly describe patterns you observe in the resulting CBRs.
  
## Answer 1 

```{r}
## calculate total person-years
Sweden$py.total <- Sweden$py.men + Sweden$py.women
Kenya$py.total <- Kenya$py.men + Kenya$py.women
World$py.total <- World$py.men + World$py.women
## calculate the CBR
Sweden.CBR <- c(sum(Sweden$births[1:15]) / sum(Sweden$py.total[1:15]), 
                sum(Sweden$births[16:30]) / sum(Sweden$py.total[16:30])) 
Kenya.CBR <- c(sum(Kenya$births[1:15]) / sum(Kenya$py.total[1:15]),
               sum(Kenya$births[16:30]) / sum(Kenya$py.total[16:30]))
World.CBR <- c(sum(World$births[1:15]) / sum(World$py.total[1:15]),
               sum(World$births[16:30]) / sum(World$py.total[16:30]))
## assign labels
names(Sweden.CBR) <- c("1950-1955", "2005-2010")
names(Kenya.CBR) <- c("1950-1955", "2005-2010")
names(World.CBR) <- c("1950-1955", "2005-2010")
## alternative, more efficient way to assign the same labels
names(Sweden.CBR) <- names(Kenya.CBR) <- names(World.CBR) <- 
    c("1950-1955", "2005-2010")
## display results
Sweden.CBR
Kenya.CBR
World.CBR
```

Since we repeat the same computation for each data frame, we can
create a function and then apply it repeatedly.  This results in a
much more efficient code.

```{r}
CBR <- function(data) {
    py.total <- data$py.men + data$py.women
    out <- c(sum(data$births[1:15]) / sum(py.total[1:15]), 
             sum(data$births[16:30]) / sum(py.total[16:30]))
    names(out) <- c("1950-1955", "2005-2010")
    return(out)
}
CBR(Sweden)
CBR(Kenya)
CBR(World)
```

We observe that during the period from 1950 to 1955, Kenya had a CBR
well above the world average whereas Sweden's was well below. Looking
at the data from 2005 to 2010, we find that the CBRs of both countries
have decreased relative to the first time period.  This appears to be
a world-wide trend.  

Note: Most countries have seen decreased fertility in the 20th and
21st centuries, though at different rates.
   
## Question 2
  
The CBR is easy to understand but contains both men and women
   of all ages in the denominator. We next calculate the *total
     fertility rate* (TFR). Unlike the CBR, the TFR adjusts for age
   compositions in the female population.  To do this, we need to
   first calculate the *age specific fertility rate* (ASFR), which
   represents the fertility rate for women of the reproductive age
   range $[15, 50)$.  The ASFR for age range $[x, x+\delta)$, where
   $x$ is the starting age and $\delta$ is the width of the age range
   (measured in years), is defined as:
  $$
    \text{ASFR}_{[x,\ x+\delta)} 
    \ = \ \frac{\text{number of births to women of age $[x,\ x+\delta)$}}{\text{Number of person-years lived by women of age $[x,\ x+\delta)$}}
  $$
  Note that square brackets, $[$ and $]$, include the limit whereas
  parentheses, $($ and $)$, exclude it.  For example, $[20, 25)$
  represents the age range that is greater than or equal to 20 years
  old and less than 25 years old.  In typical demographic data, the
  age range $\delta$ is set to 5 years.  Compute the ASFR for Sweden
  and Kenya as well as the entire world for each of the two periods.
  Store the resulting ASFRs separately for each region.  What does the
  pattern of these ASFRs say about reproduction among women in Sweden
  and Kenya?
  
## Answer 2
  
```{r}
## calculate ASFR
Sweden.ASFR <- Sweden$births / Sweden$py.women
Kenya.ASFR <- Kenya$births / Kenya$py.women
World.ASFR <- World$births / World$py.women
## add labels
names(Sweden.ASFR) <- names(Kenya.ASFR) <- names(World.ASFR) <- 
    Sweden$age
## results for women in reproductive age range [15, 50)
Sweden.ASFR[4:10]  # 1950-1955
Sweden.ASFR[19:25] # 2005-2010
Kenya.ASFR[4:10]
Kenya.ASFR[19:25]
World.ASFR[4:10] 
World.ASFR[19:25]
```

Again, we can create a function to make our code more efficient. 

```{r}
ASFR <- function(data) {
    out <- data$births / data$py.women
    names(out) <- data$age
    return(out[c(4:10, 19:25)])
}
ASFR(Sweden)
ASFR(Kenya)
ASFR(World)
```

Generally, the ASFR has declined over time for both countries and in
the entire world.  The ASFR for both Sweden and Kenya was highest for
the age range $[20, 25)$ in 1950--1955. However, in 2005--2010,
Sweden's ASFR was highest for $[30, 35)$ while Kenya's remained
highest for $[20, 25)$.  This suggests that Swedish women are delaying
childbearing until later ages compared to women in Kenya.  In the
entire world, the ASFR was highest for the age range $[20, 30)$ in
both periods. 

Note: In many developed countries like Sweden, women are delaying
childbearing until later in their lives.

## Question 3

Using the ASFR, we can define the TFR as the average number of
  children women give birth to if they live through their entire
  reproductive age.
  $$
  \text{TFR} 
   =   \text{ASFR}_{[15,\ 20)} \times 5 + \text{ASFR}_{[20,\ 25)} \times 5 
  + \dots + \text{ASFR}_{[45,\ 50)} \times 5
  $$
  
  We multiply each age-specific fertility rate rate by 5 because the
  age range is 5 years.  Compute the TFR for Sweden and Kenya as well
  as the entire world for each of the two periods.  As in the previous
  question, continue to assume that women's reproductive age range is
  $[15, 50)$.  Store the resulting two TFRs for each country or the
  world as a vector of length two. In general, how has the number of
  women changed in the world from 1950 to 2000? What about the total
  number of births in the world?

## Answer 3

```{r}
## calculate TFR for women's reproductive age range [15, 50)
Sweden.TFR <- c(sum(Sweden.ASFR[4:10] * 5), sum(Sweden.ASFR[19:25] * 5))
Kenya.TFR <- c(sum(Kenya.ASFR[4:10] * 5), sum(Kenya.ASFR[19:25] * 5))
World.TFR <- c(sum(World.ASFR[4:10] * 5), sum(World.ASFR[19:25] * 5))
## add labels
names(Sweden.TFR) <- names(Kenya.TFR) <- names(World.TFR) <-
    c("1950-1955", "2005-2010") 
## display results
Sweden.TFR 
Kenya.TFR 
World.TFR
```

We can create a function to do this computation, which calls the
`ASFR` function we created above.

```{r}
TFR <- function(data) {
    out <- c(sum(ASFR(data)[1:7] * 5), sum(ASFR(data)[8:14] * 5))
    names(out) <- c("1950-1955", "2005-2010") 
    return(out)
}
TFR(Sweden)
TFR(Kenya)
TFR(World)
```

The resulting TFRs present further evidence of the worldwide reduction
in fertility.  Women in Kenya were, on average, having
`r round(Kenya.TFR[1], digits = 1)` children in the 1950s, which
decreased to `r round(Kenya.TFR[2], digits = 1)` children in 2010.
This is a substantial decrease, but Kenya's fertility rate remains
high compared to the rest of the world.  In Sweden, we find that
fertility decreased below `r ceiling(Sweden.TFR[2]) in the second
of the two time periods.  This suggests that without immigration, the
population of Sweden may start to decrease. 

Note: In many developed countries, women are having children at later
ages and having fewer of them. World population growth is mainly
driven by relatively high fertility rates in less developed countries.

## Question 4

Next, we will examine another important demographic process:
  death.  Compute the *crude death rate* (CDR), which is a concept
  analogous to the CBR, for each period and separately for each
  region.  Store the resulting CDRs for each country and the world as
  a vector of length two. The CDR is defined as:
  $$ 
    \text{CDR} 
     =  \frac{\text{number of deaths}}{\text{number of person-years lived}}
  $$
  Briefly describe patterns you observe in the resulting CDRs.

## Answer 4

```{r}
## calculate the CDR
Sweden.CDR <- c(sum(Sweden$deaths[1:15]) / sum(Sweden$py.total[1:15]), 
                sum(Sweden$deaths[16:30]) / sum(Sweden$py.total[16:30]))
Kenya.CDR <- c(sum(Kenya$deaths[1:15]) / sum(Kenya$py.total[1:15]), 
               sum(Kenya$deaths[16:30]) / sum(Kenya$py.total[16:30]))
World.CDR <- c(sum(World$deaths[1:15]) / sum(World$py.total[1:15]), 
               sum(World$deaths[16:30]) / sum(World$py.total[16:30]))
## add labels
names(Kenya.CDR) <- names(Sweden.CDR) <- names(World.CDR) <- 
    c("1950-1955", "2005-2010")
## display results
Sweden.CDR
Kenya.CDR
World.CDR
```

As for the CBR, we can create our own function to do the computation.

```{r}
CDR <- function(data) {
    out <-  c(sum(data$deaths[1:15]) / sum(data$py.total[1:15]), 
              sum(data$deaths[16:30]) / sum(data$py.total[16:30]))
    names(out) <-  c("1950-1955", "2005-2010")
    return(out)
}
CDR(Sweden)
CDR(Kenya)
CDR(World)
```

We observe that the CDR has declined over time for Kenya and the world
while it has stayed relatively unchanged for Sweden.

## Question 5

One puzzling finding from the previous question is that the CDR
  for Kenya during the period of 2005-2010 is about the same level as
  that for Sweden.  We would expect people in developed countries like
  Sweden to have a lower death rate than those in developing countries
  like Kenya.  While it is simple and easy to understand, the CDR does
  not take into account the age composition of a population.  We
  therefore compute the *age specific death rate* (ASDR).  The
  ASDR for age range $[x, x+\delta)$ is defined as:
  $$
    \text{ASDR}_{[x,\ x+\delta)} 
    \ = \ \frac{\text{number of deaths for people of age $[x,\ x+\delta)$}}{\text{number of person-years of people of age $[x,\ x+\delta)$}}
  $$
  Calculate the ASDR for each age group, separately for Kenya and
  Sweden, during the period of 2005-2010.  Briefly describe the
  pattern you observe.
 
## Answer 5

```{r}
## compute ASDR
Sweden.ASDR <- Sweden$deaths / Sweden$py.total
Kenya.ASDR <- Kenya$deaths / Kenya$py.total
## add labels
names(Sweden.ASDR) <- names(Kenya.ASDR) <- Sweden$age
## display results in 1000 people 
Sweden.ASDR[16:30] * 1000
Kenya.ASDR[16:30] * 1000
```

If we use a function, we have the following code chunk:

```{r}
ASDR <- function(data) {
    out <- data$deaths / data$py.total
    names(out) <- data$age
    return(out[16:30] * 1000)
}
ASDR(Sweden)
ASDR(Kenya)
```

As expected, both Kenya and Sweden exhibit a U-shaped pattern. Infant
mortality, i.e., death rate for age range $[0, 5)$, is high, while
death rate drops in childhood and young adulthood.  Then, the
mortality rate begins to rise again as individuals get older. While
this general trend is observed in both countries, we see that at every
age, Sweden has a lower ASDR than Kenya.  For example, for ages [0, 5)
Kenya's ASDR is approximately
`r round(Kenya.ASDR[1]/Sweden.ASDR[1],0)` times that of Sweden.
This contrasts with the observation that the CDR during 2005-2010 for
Sweden is almost as the same as for Kenya.

## Question 6

One way to understand the difference in the CDR between Kenya
  and Sweden is to compute the counterfactual CDR for Kenya using
  Sweden's population distribution (or vice versa).  This can be done
  by applying the following alternative formula for the CDR.
  $$
    \text{CDR}
    \ = \ \text{ASDR}_{[0, 5)} \times P_{[0,5)} + \text{ASDR}_{[5, 10)}
    \times P_{[5, 10)} + \cdots 
  $$
  where $P_{[x, x+\delta)}$ is the proportion of the population in the
  age range $[x, x+\delta)$.  We compute this as the ratio of
  person-years in that age range relative to the total person-years
  across all age ranges.  To conduct this counterfactual analysis, we
  use $\text{ASDR}_{[x,x+\delta)}$ from Kenya and $P_{[x,x+\delta)}$
  from Sweden during the period of 2005--2010.  That is, first
  calculate the age-specific population proportions for Sweden and
  then use them to compute the counterfactual CDR for Kenya. How does
  this counterfactual CDR compare with the original CDR of Kenya?
  Briefly interpret the result.

## Answer 6

```{r}
## calculate population proportions
Sweden.pop <- Sweden$py.total[16:30] / sum(Sweden$py.total[16:30])
Kenya.pop <- Kenya$py.total[16:30] / sum(Kenya$py.total[16:30])
## add labels
names(Sweden.pop) <- Sweden$age[16:30]
names(Kenya.pop) <- Kenya$age[16:30]
## compare the two population distributions
Sweden.pop
Kenya.pop
## Kenyan CDR with Swedish population proportions
sum(Kenya.ASDR[16:30] * Sweden.pop)
## compare it with the original CDRs 
Kenya.CDR 
Sweden.CDR
```

The counterfactual CDR for Kenya is nowhere near as low as the
original CDR for Kenya and is much higher than the CDR for Sweden.
Because Sweden's population is much older than Kenya's counterpart
(and older people have higher mortality rates), the CDR for Sweden is
about the same level as that for Kenya even though age-specific death
rates are much lower for the Swedish population.  The analysis shows
how much the age composition of a population can influence the CDR.
The difference in the CDR can come from the difference in population
distribution as well as the difference in death rates.

