---
title: "Oil, Democracy, and Development"
output:
  pdf_document: default
  html_document: default
---

Researchers have theorized that natural resources may have an
inhibiting effect on the democratization process.  Although 
there are multiple explanations as to why
this might be the case, one hypothesis posits that governments in
countries with large natural resource endowments (like oil) are able
to fund their operations without taxing civilians. Since
representation (and other democratic institutions) are a compromise
offered by governments in exchange for tax revenue, resource-rich
countries do not need to make this trade.  In this exercise, we will
not investigate causal effects of oil on democracy.  Instead, we
examine whether the association between oil and democracy is
consistent with the aforementioned hypothesis.  

This exercise is in part based on Michael L. Ross. (2001). 'Does Oil
 Hinder Democracy?' *World Politics*,53:3,
 pp.325-361.

The data set is in the
csv file `resources.csv`.  The names and descriptions of variables are:


-------------------------------------------------------------------------------
 Name                 Description
 -------------------- ---------------------------------------------------------
 `cty_name`           Country name
 
 `year`               Year
 
 `logGDPcp`           Logged GDP per capita 
 
 `regime`             A measure of a country's level of democracy: -10 (authoritarian) 
                      to 10 (democratic) 
 
 `oil`                Amount of oil exports as a percentage of the country's GDP 
 
 `metal`              Amount of non-fuel mineral exports as a percentage of the country's GDP 
 
 `illit`              Percentage of the population that is illiterate 
 
 `life`               Life expectancy in the country 
-------------------------------------------------------------------------------

## Question 1

Use scatterplots to examine the bivariate relationship between
  logged GDP per capita and life expectancy as well as between logged
  GDP per capita and illiteracy.  Be sure to add informative axis
  labels. Also, compute the correlation separately for each bivariate
  relationship.  Briefly comment on the results.  To remove missing
  data when applying the `cor` function, set `use` argument
  to `"complete.obs"`.
  
## Answer 1  
  
```{r}
resources <- read.csv("data/resources.csv")

par(cex = 1.25)
plot(x = resources$logGDPcp, y = resources$life,
     ylab = "Life Expectancy", xlab = "Logged GDP Per Capita")
plot(x = resources$logGDPcp, y = resources$illit,
     ylab = "Illiteracy Rate", xlab = "Logged GDP Per Capita")
cor(resources$logGDPcp, resources$life, use = "complete.obs")
cor(resources$logGDPcp, resources$illit, use = "complete.obs")
```

As we might expect, higher levels of GDP per capita are associated
with higher life expectancy and lower illiteracy rates.

## Question 2

We focus on the following subset of the variables:
`regime`, `oil`, `logGDPcp`, and
  `illit`. Remove observations that have missing values in any
  of these variables.  Using the `scale()` function, scale these
  variables so that each variable has a mean of zero and a standard
  deviation of one.  Fit the k-means clustering algorithm with two
  clusters.  How many observations are assigned to each cluster?  Using
  the original unstandardized data, compute the means of these variables in each cluster.


## Answer 2

```{r}
res <- subset(resources, select = c("cty_name", "year", "regime",
                             "oil", "logGDPcp", "illit"))
resNoNA <- na.omit(res)
resNoNAS <- scale(resNoNA[, 3:6])
colMeans(resNoNA[, 3:6])
apply(resNoNA[, 3:6], 2, sd)
cluster2 <- kmeans(resNoNAS, centers = 2)
table(cluster2$cluster)
## this gives standardized answer, which is hard to interpret
cluster2$centers
## better to subset the original data and then compute means
g1 <- resNoNA[cluster2$cluster == 1, ]
colMeans(g1[, 3:6])
g2 <- resNoNA[cluster2$cluster == 2, ]
colMeans(g2[, 3:6])
```
  
Countries assigned to the first cluster tend to have higher levels of
oil but also lower levels of democracy, lower GDP per capita, and
higher rates of illiteracy. This finding aligns with the claim that
higher levels of natural resources may interfere with democratization.

## Question 3 

Using the clusters obtained above, modify the scatterplot
  between logged GDP per capita and illiteracy rate in the following
  manner.  Use different colors for the clusters so that we can easily
  tell the cluster membership of each observation.  In addition, make
  the size of each circle proportional to the `oil` variable so
  that oil-rich countries stand out.  Briefly comment on the results.

## Answer 3

```{r}
plot(x = resNoNA$logGDPcp, y = resNoNA$illit,
     col = cluster2$cluster, cex = resNoNA$oil / 10 + 1,
     ylab = "Illiteracy Rate", xlab = "Logged GDP Per Capita")  
```

The figure makes evident that countries clustered into the `resource
rich' cluster typically have lower literacy rates and lower per
capita GDPs. Importantly, the figure shows that many of the countries
with relatively high oil exports also tend to have relatively high per
capita GDPs.

## Question 4

Repeat the previous two questions but this time with three
  clusters instead of two.  How are the results different?  Which
  clustering model would you prefer and why?

## Answer 4

```{r}
set.seed(1234)
cluster3 <- kmeans(resNoNAS, centers = 3)
g1 <- resNoNA[cluster3$cluster == 1, ]
colMeans(g1[, 3:6])
g2 <- resNoNA[cluster3$cluster == 2, ]
colMeans(g2[, 3:6])
g3 <- resNoNA[cluster3$cluster == 3, ]
colMeans(g3[, 3:6])
plot(x = resNoNA$logGDPcp, y = resNoNA$illit,
     col = cluster3$cluster, cex = resNoNA$oil / 10 + 1,
     ylab = "Illiteracy Rate", xlab = "logged GDP per capita") 
```

The revised model distinguishes the largest oil exporters from less
democratized countries with low oil exports. This model is preferable
because provides greater precision for characterizing oil rich
countries. Compared to countries with relatively low oil exports, we
find that the largest oil exporters tend to have more authoritarian
regimes. Moreover, these countries tend to have relatively high per
capita GDPs but also relatively high illiteracy rates.

