---
title: "Indiscriminate Violence and Insurgency"
output: pdf_document
---

In this exercise, we analyze the relationship between indiscriminate
violence and insurgent attacks using data about Russian artillery fire
in Chechnya from 2000 to 2005. 

This exercise is based on: Lyall, J. 2009. “[Does Indiscriminate Violence 
Incite Insurgent Attacks?: Evidence from 
Chechnya.](http://dx.doi.org/10.1177/0022002708330881)” 
 *Journal of Conflict Resolution* 53(3): 331–62.
    
Some believe that
indiscriminate violence increases insurgent attacks by creating more
cooperative relationships between citizens and insurgents.  Others
believe that indiscriminate violence can be effective in suppressing
insurgent activities.  

This dataset was constructed around 159 events in which Russian
artillery shelled a village.  For each such event we record 
the village where the shelling took place and whether it was in Groznyy, 
how many people were killed, and the number of 
insurgent attacks 90 days before and 90 days after the date of the event.
We then augment this data by observing the same information 
for a set of demographically and geographically
similar villages that were not shelled during the same time periods.

The names and
descriptions of variables in the data file `chechen.csv` are

--------------------------------------------------------------------------------
 Name                 Description
 -------------------- ----------------------------------------------------------
 `village`            Name of village
 
 `groznyy`            Variable indicating whether a village is in
                      Groznyy (1) or not (0) 
                      
 `fire`               Whether Russians struck a village with artillery fire 
                      (1) or not (0)
                      
 `deaths`             Estimated number of individuals killed during Russian artillery fire 
                      or NA if not fired on
                      
 `preattack`          The number of insurgent attacks in the 90 days before being fired on
 
 `postattack`         The number of insurgent attacks in the 90 days after being fired on 
--------------------------------------------------------------------------------

Note that the same village may
appear in the dataset several times as shelled and/or not shelled
because Russian attacks occurred at different times and locations.

## Question 1
  
How many villages were shelled by Russians?  How many were not?
  
## Answer 1 
```{r}
chechen <- read.csv("data/chechen.csv")

# Villages ever attacked
attacked <- length(unique(chechen$village[chechen$fire==1]))
attacked
# All villages
total <- length(unique(chechen$village))
# Villages never attacked
total - attacked
```

`r attacked` villages were shelled, often more than once, and `r total - attacked` villages were never attacked.

## Question 2

Were artillery attacks on
  Groznyy more lethal than attacks on villages outside of Groznyy?  
  Conduct the
  comparison in terms of the mean and median.

## Answer 2   

```{r}
# Note: deaths is NA for rows with no attack
tapply(chechen$deaths, chechen$groznyy, mean, na.rm = TRUE)
tapply(chechen$deaths, chechen$groznyy, median, na.rm = TRUE)
```
   
Yes, artillery attacks killed on average
`r round(tapply(chechen$deaths, chechen$groznyy, mean,
  na.rm=TRUE), 1)[[2]]` people in Groznyy but only
`r round(tapply(chechen$deaths, chechen$groznyy, mean,
  na.rm=TRUE), 2)[[1]]` people from villages outside Grozny.

## Question 3
    
Compare the average number of insurgent attacks for observations
describing a shelled village and the others.
Also, compare the quartiles. Would you conclude that indiscriminate
violence reduces insurgent attacks?  Why or why not?
    
## Answer 3

```{r}
tapply(chechen$postattack, chechen$fire, mean)
tapply(chechen$postattack, chechen$fire, quantile)
```

The quartiles reveal that the distribution of insurgent attacks 
in periods after shelling are similar between the villages 
hit by artillery fire and those that were not.  The average number of 
insurgent attacks is
`r (round(tapply(chechen$postattack, chechen$fire, mean)[[2]],2))`
for observations of villages that were shelled vs 
`r (round(tapply(chechen$postattack, chechen$fire, mean)[[1]],2))`
for the others.  By itself, this comparison
suggests that indiscriminate violence may slightly reduce insurgent attacks.
 
## Question 4

Considering only the pre-shelling periods, what is the difference between the average number of insurgent attacks for observations describing a shelled village and observations that do not?   What does this suggest to 
you about the validity of comparison used for the previous question?

## Answer 4

```{r}
av_pre <- tapply(chechen$preattack, chechen$fire, mean)
av_pre[2] - av_pre[1]
```

In these periods the average number of insurgency attacks was similar but slightly lower in observations of villages that were later shelled.  
Although this is an observational study where artillery fire is not 
formally randomized, the similarity of these averages
increases the credibility of the comparison of insurgency attacks
between villages hit by Russian fire and those not.
 
## Question 5   
    
Create a new variable called `diffattack` by calculating
  the difference in the number of insurgent attacks in the before 
  and after periods.  Among observations describing villages that were shelled
  did the number of insurgent attacks increase after being fired on?  
  Give a substantive interpretation of the result.

## Answer 5

```{r}
# Effect: post-treatment - pre-treatment
chechen$diffattack <- chechen$postattack - chechen$preattack
# Proportion of villages that had more attacks after than before
attack_increase <- tapply(chechen$diffattack > 0, chechen$fire, mean) 
attack_increase
```

`r round(attack_increase[2]*100, 2)` percent of Russian artillery attacks
were followed by an increase in insurgent attacks, 
compared to `r round(attack_increase[1]*100, 2)` percent for observations
where there was no attack.

## Question 6

Compute the mean difference in the `diffattack` variable
  between observations where villages were shelled and those where 
  they were not.  Does this
  analysis support the claim that indiscriminate violence reduces
  insurgency attacks?  Is the validity of this analysis improved over
  the analyses you conducted in the previous questions?  Why or why
  not?  Specifically, explain what additional factor this analysis
  addresses when compared to the analyses conducted in the previous
  questions.

## Answer 6

```{r}
mean.diff <- tapply(chechen$diffattack, chechen$fire, mean)
mean.diff[2] - mean.diff[1]
```

The results support the conclusion that indiscriminate
violence reduces insurgent attacks.  This analysis takes into
account the common time trend that exists for the two types of
observations.  If we assume that the trend in numbers of attacks
pre- and post-shelling among villages that were not in fact shelled 
is what we would have observed in shelled villages had they not been shelled, then the difference in the
differences between pre- and post-shelling attacks between the two
village types can be attributed to the Russian artillery fire.
