---
title: "Ideology of US Supreme Court Justices"
output:
  pdf_document: default
  html_document: default
---

We introduced an important programming method called the *loop*.
In this exercise, we practice using loops 
with data on the ideological positions of United States
Supreme Court Justices.  Just like legislators, justices make voting decisions 
that we can use to estimate their ideological positions.  This exercise
  is based in part on Andrew Martin and Kevin Quinn. (2002). 'Dynamic
  Ideal Point Estimation via Markov Chain Monte Carlo for the
  U.S. Supreme Court, 1953-1999.' *Political Analysis*, 10:2, pp.134-154.
  
The file `justices.csv` contains the following variables:

--------------------------------------------------------------------------------
 Name                 Description
 -------------------- ---------------------------------------------------------
 `term`               Supreme Court term (a year)
 
 `justice`            Justice's name 
 
 `idealpt`            Justice's estimated ideal point in that term
 
 `pparty`             Political party of the president in that term
 
 `pres`               President's name
--------------------------------------------------------------------------------

The ideal points of the justices are negative to indicate liberal 
preferences and positive to indicate conservative preferences. 

## Question 1

We wish to know the median ideal point for the Court during each
  term included in the dataset. First, calculate the median ideal point
  for each term of the Court. Next, generate a plot
  with term on the horizontal axis and ideal point on the vertical axis. 
  Include a dashed horizontal line at zero to indicate a "neutral"
  ideal point. Be sure to include informative axis labels
  and a plot title.

## Answer 1

```{r}
justices <- read.csv("data/justices.csv", stringsAsFactors = FALSE)
median_ip <- tapply(justices$idealpt, justices$term, median)
```

```{r}
terms <- unique(justices$term)
plot(terms, median_ip, xlab = "US Supreme Court Term", 
     ylab = "Median Ideal Point", type = "b")
abline(h = 0, col = "blue", lty = "dashed")
text(1995, 0, "Neutral Ideal Point", pos = 1)
```

## Question 2

Next, we wish to identify the name of the justice
  with the median ideal point **for each term**. Which justice had the median
  ideal point in the most (potentially
  nonconsecutive) terms? How long did this justice serve on the Court overall?
  What was this justice's average ideal point over his/her entire
  tenure on the Court?

## Answer 2

```{r}
median_justice <- rep(NA, length(terms))
names(median_justice) <- terms
for (i in 1:length(terms)){
  med <- justices$term == terms[i] & justices$idealpt == median_ip[i]
  median_justice[i] <- justices$justice[med]
}
table(median_justice) # White is median 13 times

white <- subset(justices, justice == "White")
white_tenure <- nrow(white)
white_average_ip <- mean(white$idealpt) 
```

White served as a justice for `r white_tenure` years, of which he
was the median justice for `r table(median_justice)["White"]` years
- the most ever in the postwar period. His mean ideal point was
`r round(white_average_ip, 2)`, indicating a
moderate-conservative leaning.

## Question 3

We now turn to the relationship between Supreme Court
  ideology and the president. Specifically, we want to see how
  the ideology of the Supreme Court changes over the course of each president's
  time in office. Begin by creating two empty
  'container' vectors: one to hold Democratic presidents, 
  and another for Republican presidents. Label each vector with the presidents' names. 
  
## Answer 3

```{r}
dem_pres_names <- unique(justices$pres[justices$pparty == "D"])
dem_pres_shift <- rep(NA, length(dem_pres_names))
names(dem_pres_shift) <- dem_pres_names

rep_pres_names <- unique(justices$pres[justices$pparty == "R"])
rep_pres_shift <- rep(NA, length(rep_pres_names))
names(rep_pres_shift) <- rep_pres_names
```

## Question 4

Next, for each Democratic president, calculate the shift in Supreme
  Court ideology by subtracting the Court's median ideal point in the 
  president's first term from its median ideal point in the president's last
  term. Use a loop to store these values in your Democratic container vector. 
  Repeat the same process for Republican presidents. 

## Answer 4

```{r}
## Democrats
for (pres_name in dem_pres_names) {
  pres <- subset(justices, pres == pres_name)
  med_ip <- tapply(pres$idealpt, pres$term, median)
  dem_pres_shift[pres_name] <- med_ip[length(med_ip)] - med_ip[1]
}

for (pres_name in rep_pres_names) {
  pres <- subset(justices, pres == pres_name)
  med_ip <- tapply(pres$idealpt, pres$term, median)
  rep_pres_shift[pres_name] <- med_ip[length(med_ip)] - med_ip[1]
}
```

## Question 5

What was the mean and standard deviation of the Supreme Court
  ideology shifts you just calculated when looking only at the 
  Democratic presidencies? What about the Republican presidencies?
  Which Republican president's tenure had the largest conservative 
  (positive) shift on the Court? Which Democratic president's tenure
  had the largest liberal (negative) shift? 

## Answer 5

```{r}
mean(dem_pres_shift)
sd(dem_pres_shift)

mean(rep_pres_shift)
sd(rep_pres_shift)

dem_pres_shift[dem_pres_shift == min(dem_pres_shift)]
rep_pres_shift[rep_pres_shift == max(rep_pres_shift)]
```

## Question 6

Create a plot that shows the median Supreme Court ideal point
  over time.  Then, add lines for the ideal points of each unique justice
  to the same plot. The color of each line should be red if the 
  justice was appointed by a Republican and blue if he or she was appointed
  by a Democrat. (You can assume that when a Justice first appears in the 
  data, they were appointed by the president sitting during that term.)
  Label each line with the justice's last name.  Briefly comment on the
  resulting plot.

## Answer 6

```{r}
justice_names <- unique(justices$justice)

plot(terms, median_ip,
     type = "l", xlim = c(1948, 2015), ylim = c(-7, 7), 
     xlab = "Year", ylab = "Ideology", lwd = 4, col='grey',
     main = "Supreme Court Justice Ideological Movement")

abline(h = 0, lty = "dotted", col = "grey")

for (name in justice_names) {
    just <- subset(justices, justices$justice == name)
    partycol <- ifelse(just$pparty[1] == "R", "red", "blue")
    lines(just$term, just$idealpt, col = partycol)
    last_term <- nrow(just)
    text(just$term[last_term], just$idealpt[last_term],
         label = name, pos = 4)
}

```

In general, we find that most Supreme Court justices shift toward more
liberal ideologies. It is also worth noting that while ideologies of
Supreme court justices in the mid-twentieth century were relatively
similar, we find a much larger range of ideology in recent years. This
is particularly interesting since in 2010, most Supreme Court justices
were nominees of Republican Presidents.
