---
title: "Election and Conditional Cash Transfer Program in Mexico"
output:
  pdf_document: default
  html_document: default
date: "5 August 2015"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(error=FALSE,comment='')
```

In this exercise, we analyze the data from a study that seeks to
estimate the electoral impact of 'Progresa', Mexico's
*conditional cash transfer program* (CCT program). This
  exercise is based on the following article: Ana de la O. (2013).
  'Do Conditional Cash Transfers Affect Voting Behavior? Evidence
  from a Randomized Experiment in Mexico.'  *American Journal of
    Political Science*, 57:1, pp.1-14. and Kosuke Imai,
  Gary King, and Carlos Velasco. (2015). 'Do Nonpartisan Programmatic
  Policies Have Partisan Electoral Effects? Evidence from Two Large
  Scale Randomized Experiments.' Working Paper.
  
The original study
relied on a randomized evaluation of the CCT program in which eligible
villages were randomly assigned to receive the program either 21
(Early *Progresa*) or 6 months (Late *Progresa*) before
the 2000 Mexican presidential election. The author of the original
study hypothesized that the CCT program would mobilize voters, leading
to an increase in turnout and support for the incumbent party (PRI in
this case).  The analysis was based on a sample of precincts that
contain at most one participating village in the evaluation.

The data we analyze are available as the CSV file
`progresa.csv`. The names
and descriptions of variables in the data set are:

--------------------------------------------------------------------------------
 Name                 Description
 -------------------- ----------------------------------------------------------
 `treatment`          Whether an electoral precinct contains a village where 
                      households received Early *Progresa*
                      
 `pri2000s`           PRI votes in the 2000 election as a share of precinct 
                      population above 18
 
 `pri2000v`           Official PRI vote share in the 2000 election
 
 `t2000`              Turnout in the 2000 election as a share of precinct 
                      population above 18
                      
 `t2000r`             Official turnout in the 2000 election
 
 `pri1994`            Total PRI votes in the 1994 presidential election
 
 `pan1994`            Total PAN votes in the 1994 presidential election 
 
 `prd1994`            Total PRD votes in the 1994 presidential election
 
 `pri1994s`           Total PRI votes in the 1994 election as a share of 
                      precinct population above 18
                      
 `pan1994s`           Total PAN votes in the 1994 election as a share of 
                      precinct population above 18
                      
 `prd1994s`           Total PRD votes in the 1994 election as a share of 
                      precinct population above 18
                      
 `pri1994v`           Official PRI vote share in the 1994 election
 
 `pan1994v`           Official PAN vote share in the 1994 election
 
 `prd1994v`           Official PRD vote share in the 1994 election
 
 `t1994`              Turnout in the 1994 election as a share of precinct 
                      population above 18
                      
 `t1994r`             Official turnout in the 1994 election
 
 `votos1994`          Total votes cast in the 1994 presidential election 
 
 `avgpoverty`         Precinct Avg of Village Poverty Index
 
 `pobtot1994`         Total Population in the precinct
 
 `villages`           Number of villages in the precinct
---------------------------------------------------------------------------------

Each observation in the
data represents a precinct, and for each precinct the file contains
information about its treatment status, the outcomes of interest,
socioeconomic indicators, and other precinct characteristics.

## Question 1

Estimate the impact of the CCT program on turnout and support
  for the incumbent party (PRI or Partido Revolucionario
  Institucional) by comparing the average electoral outcomes in the
  'treated' (Early *Progresa*) precincts versus the ones
  observed in 'control' (Late *Progresa*) precincts. Next,
  estimate these effects by regressing the outcome variable on the
  treatment variable. Interpret and compare the estimates under these
  approaches.  Here, following the original analysis, use the turnout
  and support rates as shares of the voting eligible population
  (`t2000` and `pri2000s`, respectively).  Do the results
  support the hypothesis?  Provide a brief interpretation.

## Answer 1

```{r}
cct <- read.csv("data/progresa.csv")
## effect on turnout under standard difference-in-means
itt.turn <- mean(cct$t2000[cct$treatment == 1]) - 
    mean(cct$t2000[cct$treatment == 0])
itt.turn
## effect on turnout under linear regression
ols.turn <- lm(t2000 ~ treatment, data = cct)
ols.turn
## effect on PRI support under standard difference-in-means
itt.pri <- mean(cct$pri2000s[cct$treatment == 1]) - 
    mean(cct$pri2000s[cct$treatment == 0])
itt.pri
## effect on PRI support under linear regression
ols.pri <- lm(pri2000s ~ treatment, data = cct)
ols.pri
```

We find that the estimated average effect of the CCT on turnout and
support for the incumbent party are `r round(itt.turn, 2)` and
`r round(itt.pri, 2)` percentage points, respectively.  As we saw
earlier in the chapter, these estimates are identical regardless of
whether one uses the difference-in-means or linear regression
estimators.  Each estimate should be interpreted as the impact of *Progresa*'s availability rather than the receipt of the program
itself.  The results are consistent with the hypothesis, suggesting
that on average, the availability of the CCT program increased turnout
and support for the incumbent party by about 4 percentage points.

## Question 2

In the original analysis, the authors fit a linear regression
  model that includes, as predictors a set of pre-treatment covariates
  as well as the treatment variable.  Here, we fit a similar model for
  each outcome that includes the average poverty level in a precinct
  (`avgpoverty`), the total precinct population in 1994
  (`pobtot1994`), the total number of voters who turned out in
  the previous election (`votos1994`), and the total number of
  votes cast for each of the three main competing parties in the
  previous election (`pri1994` for PRI, `pan1994` for
  Partido Acción Nacional or PAN, and `prd1994` for Partido
  de la Revolución Democrática or PRD).  Use the same outcome
  variables as in the original analysis that are based on the shares
  of the voting age population.  According to this model, what are the
  estimated average effects of the program's availability on turnout
  and support for the incumbent party?  Are these results different
  from what you obtained in the previous question?
  
## Answer 2

```{r}
## effect on turnout under original analysis
org.turn <- 
    lm(t2000 ~ treatment + avgpoverty + pobtot1994 + 
           votos1994 + pri1994 + pan1994 + prd1994, data = cct)
org.turn
## effect on PRI support under original analysis
org.pri <- 
    lm(pri2000s ~ treatment + avgpoverty + pobtot1994 +
           votos1994 + pri1994 + pan1994 + prd1994, data = cct)
org.pri
```

The intention-to-treat estimates of *Progresa* on turnout and
support for the incumbent party under the original regression analysis
are now `r round(coef(org.turn)[2], 2)` and
`r round(coef(org.pri)[2], 2)` percentage points, respectively.
The differences between the original regression-based estimates and
these difference-in-means estimates are `r round(coef(org.turn)[2]
  - itt.turn, digits = 2)` and `r round(coef(org.pri)[2] -
  itt.pri, digits = 2)` percentage points for turnout and PRI support,
respectively.  These estimates are quite similar.

## Question 3

Next, we consider an alternative, and more natural, model
  specification.  We will use the original outcome variables as in the
  previous question.  However, our model should include the previous
  election outcome variables measured as shares of the voting age
  population (as done for the outcome variables `t1994`,
  `pri1994s`, `pan1994s`, and `prd1994s`) instead of
  those measured in counts. In addition, we apply the natural
  logarithm transformation to the precinct population variable when
  including it as a predictor.  As in the original model, our model
  includes the average poverty index as an additional predictor.  Are
  the results based on these new model specifications different from
  what we obtained in the previous question?  If the results are
  different, which model fits the data better?
  
## Answer 3

```{r}
## effect on Turnout (previous outcome in ratio; log population)
turn.ratio <- 
    lm(t2000 ~ treatment + avgpoverty + I(log(pobtot1994)) + 
           t1994 + pri1994s + pan1994s + prd1994s, data = cct)
turn.ratio
## effect on PRI support (previous outcome in ratio; log population)
pri.ratio <- 
    lm(pri2000s ~ treatment + avgpoverty + I(log(pobtot1994)) + 
           t1994 + pri1994s + pan1994s + prd1994s,
       data = cct)
pri.ratio
```

When we adjust for the previous outcomes in ratios instead of counts
and log-transform the precinct population, the estimates for the ITT
effect on turnout and incumbent support are
`r round(coef(turn.ratio)[2], 2)` and
`r round(coef(pri.ratio)[2], 2)` percentage points,
respectively. This result is a significant change relative to the
findings of the original analysis and suggests that program
availability has little effect on election outcomes.

We examine the model fit under each specification using the adjusted
$R^2$. We use the `adjR2` function created earlier in the
chapter.

```{r}
## adjusted Rsquare
adjR2 <- function(fit) {
    resid <- resid(fit)
    y <- fitted(fit) + resid
    n <- length(y)
    TSS.adj <- sum((y - mean(y))^2) / (n - 1)
    SSR.adj <- sum(resid^2) / (n - length(coef(fit)))
    R2.adj <- 1 - SSR.adj / TSS.adj
    return(R2.adj)
}
## original turnout model
adjR2(org.turn)
## turnout model with previous turnout in ratio
adjR2(turn.ratio)
## original PRI support model 
adjR2(org.pri)
## PRI support model with previous support for parties in ratios
adjR2(pri.ratio)
```

The model fit under the original specification is quite poor for both
turnout and PRI support (`r round(adjR2(org.turn), 2)` and
`r round(adjR2(org.pri), 2)`, respectively). When instead we
adjust for the previous outcomes in ratios and log-transform the
precinct population variable, the fit of each model improves rather
dramatically (to `r round(adjR2(turn.ratio), 2)` and
`r round(adjR2(pri.ratio), 2)`, respectively).  The poor model fit
in the original analysis may be explained by the fact that the counts
of the previous outcomes are noisy predictors.  It is also possible
that the linearity assumption may be less appropriate in the case of
counts as compared to ratios if counts contain large outliers.


## Question 4

We examine the balance of some pre-treatment variables used in
  the previous analyses.  Using boxplots, compare the distributions of
  the precinct population (on the original scale), average poverty
  index, previous turnout rate (as a share of the voting age
  population), and previous PRI support rate (as a share of the voting
  age population) between the treatment and control groups.  Comment
  on the patterns you observe.
  
## Answer 4

We begin by creating a boxplot for each of the two variables,
comparing their distributions between the treatment and control
groups.  

```{r}
par(cex = 1.5)
boxplot(cct$avgpoverty ~ cct$treatment, xlab = "Treatment status",
        ylab = "Poverty Index", main = "Poverty")
boxplot(cct$pobtot1994 / 1000 ~ cct$treatment, 
        xlab = "Treatment status", ylab = "Population (thousands)", 
        main = "Population")
boxplot(cct$t1994 ~ cct$treatment, ylim = c(0, 300),
        xlab = "Treatment status", ylab = "Turnout Rate (percent)", 
        main = "Previous Turnout") 
boxplot(cct$pri1994s ~ cct$treatment, ylim = c(0, 300),
        xlab = "Treatment status", ylab = "PRI Support Rate (percent)", 
        main = "Previous PRI Support")
```

The boxplots show that while the average poverty index is balanced
across the two groups, the population variable includes two outliers.
Both of these observations belong to the treatment group.  Similarly,
the previous turnout rate and previous PRI support rate variables are
imbalanced.  For both variables, the treatment group has a few
outliers.  In theory, such an imbalance can result by chance even when
the treatment assignment is properly randomized.  However, it could
also arise from the failure of randomization.  In any event, this
imbalance may explain why the estimated average treatment effect based
on this specific sample is sensitive to the model specification.

Furthermore, we observe that several precincts have turnout rates
greater than 100 percent.  Similarly, the PRI support rate for some
precincts exceeds 100 percent.  There appear to be significant
undercounts of eligible voters, resulting in the measurement error of
the turnout rate and PRI support rate variables.  To see if this
measurement error problem exists in the outcome variables, we
summarize the distribution of the turnout rate and PRI support rate
variables.  The boxplots show that the same exact measurement error
problem exists in the outcome variables where several observations,
mostly in the treatment group, exceed 100 percent.


```{r}
par(cex = 1.5)
boxplot(cct$t2000 ~ cct$treatment, ylim = c(0, 300),
        xlab = "Treatment status", ylab = "Turnout Rate (percent)", 
        main = "Turnout") 
boxplot(cct$pri2000s ~ cct$treatment, ylim = c(0, 300),
        xlab = "Treatment status", ylab = "PRI Support Rate (percent)", 
        main = "PRI Support") 
```


## Question 5

We next use the official turnout rate `t2000r` (as a share
  of the registered voters) as the outcome variable rather than the
  turnout rate used in the original analysis (as a share of the voting
  age population). Similarly, we use the official PRI's vote share
  `pri2000v` (as a share of all votes cast) rather than the
  PRI's support rate (as a share of the voting age population).
  Compute the average treatment effect of the CCT program using a
  linear regression with the average poverty index, the
  log-transformed precinct population, and the previous official
  election outcome variables (`t1994r` for the previous turnout;
  `pri1994v`, `pan1994v`, and `prd1994v` for the
  previous PRI, PAN, and PRD vote shares).  Briefly interpret the
  results.
  
## Answer 5

```{r}
## effects on turnout
off.turn <- 
    lm(t2000r ~ treatment + avgpoverty + I(log(pobtot1994)) + 
           t1994r + pri1994v + pan1994v + prd1994v, data = cct)
off.turn
## effects on PRI support
off.pri <- 
    lm(pri2000v ~ treatment + avgpoverty + I(log(pobtot1994)) + 
           t1994r + pri1994v + pan1994v + prd1994v, data = cct)
off.pri
```

The intention-to-treat estimates of *Progresa* on official turnout
and support for the incumbent party are
`r round(coef(off.turn)[2], 2)` and `r round(coef(off.pri)[2],
  2)`.  These estimates are rather small and do not provide strong
support for the hypothesis that the CCT program increases turnout and
support for the incumbent party.

## Question 6

So far we have focused on estimating the average treatment
  effects of the CCT program. However, these effects may vary from one
  precinct to another. One important dimension to consider is poverty.
  We may hypothesize that since individuals in precincts with higher
  levels of poverty are more receptive to the cash transfers, they are
  more likely to turn out in the election and support the incumbent
  party when receiving the CCT program.  Assess this possibility by
  examining how the average treatment effect of the policy varies by
  different levels of poverty for precincts. To do so, fit a linear
  regression with the following predictors: the treatment variable,
  the log transformed precinct population, the average poverty index
  and its square, the interaction between the treatment and poverty
  index, and the interaction between the treatment and squared poverty
  index.  Estimate the average effects for unique observed values and
  plot them as a function of the average poverty level.  Comment on
  the resulting plot.
  

## Answer 6

```{r}
par(cex = 1.5)
## unique values, sorted
poverty.unique <- sort(unique(cct$avgpoverty))
## effect on turnout with non-linear interactions 
turn.by.poverty <- 
    lm(t2000r ~ treatment + I(log(pobtot1994)) + avgpoverty + 
           I(avgpoverty^2) + avgpoverty:treatment + 
               I(avgpoverty^2):treatment, data = cct)
turn.by.poverty
## predicting turnout under treatment condition
turn.cct.hatT <- 
    predict(turn.by.poverty, newdata = data.frame(treatment = 1, 
                                 pobtot1994 = mean(cct$pobtot1994), 
                                 avgpoverty = poverty.unique))
## predicting turnout under control condition
turn.cct.hatC <- predict(turn.by.poverty, 
                         newdata = data.frame(treatment = 0, 
                             pobtot1994 = mean(cct$pobtot1994), 
                             avgpoverty = poverty.unique))
## effect on turnout with non-linear interactions 
pri.by.poverty <- 
    lm(pri2000v ~ treatment + I(log(pobtot1994)) + avgpoverty + 
           I(avgpoverty^2) + avgpoverty:treatment + 
               I(avgpoverty^2):treatment, data = cct)
pri.by.poverty
## predicted PRI support under treatment condition
pri.cct.hatT <- 
    predict(pri.by.poverty, newdata = data.frame(treatment = 1, 
                                pobtot1994 = mean(cct$pobtot1994), 
                                avgpoverty = poverty.unique))
## predicting PRI support under control condition
pri.cct.hatC <-  
    predict(pri.by.poverty, newdata = data.frame(treatment = 0, 
                                pobtot1994 = mean(cct$pobtot1994), 
                                avgpoverty = poverty.unique))
## plot average effect as a function of average poverty index
plot(poverty.unique, turn.cct.hatT - turn.cct.hatC, type = "l", 
     ylim = c(-5, 18), xlab = "Average Poverty Index",
     ylab = "Percentage Points")
lines(poverty.unique, pri.cct.hatT - pri.cct.hatC, lty = "dashed")
text(4, 6, "Average effect on PRI support")
text(4, -3, "Average effect on turnout")
```

The average effect of *Progresa*'s availability on incumbent
support appears heterogeneous. Interestingly, the estimated effect is
greater among precincts with lower values of poverty level.  In
contrast, we find little non-linearity in the estimated effect on
turnout.  Across the lower range of precinct poverty level, the
estimated effect is slightly negative, while across the upper range,
the estimated effect is slightly positive.
