---
title: "Government Transfer and Poverty Reduction in Brazil"
output: 
  html_document
---

This exercise is based on
Litschig, Stephan, and Kevin M Morrison. 2013. “[The Impact of Intergovernmental Transfers on Education Outcomes and Poverty Reduction.](http://dx.doi.org/10.1257/app.5.4.206)” *American Economic Journal: Applied Economics* 5(4): 206–40. 

In this exercise, we estimate the effects of increased government
spending on educational attainment, literacy, and poverty
rates.

Some
scholars argue that government spending accomplishes very little in
environments of high corruption and inequality. Others suggest that in
such environments, accountability pressures and the large demand for
public goods will drive elites to respond.  To address this debate, we
exploit the fact that until 1991, the formula for government transfers
to individual Brazilian municipalities was determined in part by the
municipality's population. This meant that municipalities with
populations below the official cutoff did not receive additional
revenue, while states above the cutoff did.  The data set
`transfer.csv` contains the variables:

--------------------------------------------------------------------------------
 Name                Description
 ------------------- ---------------------------------------------------------
 `pop82`             Population in 1982
 
 `poverty80`         Poverty rate of state in 1980
 
 `poverty91`         Poverty rate of state in 1991
 
 `educ80`            Average years education of state in 1980
 
 `educ91`            Average years education of state in 1991
 
 `literate91`        Literacy rate of state in 1991
 
 `state`             State
 
 `region`            Region
 
 `id`                Municipal ID
 
 `year`              Year of measurement
--------------------------------------------------------------------------------

## Question 1

We will apply the regression discontinuity design to this
  application.  State the required assumption for this design and
  interpret it in the context of this specific application.  What
  would be a scenario in which this assumption is violated?  What are
  the advantages and disadvantages of this design for this specific
  application?

## Answer 1

The main identifying assumption behind this design is that the 
pre-treatment characteristics of those cities right below the population 
cutoff do not differ discontinuously from the characteristics of the 
cities that are right above the cutoff. In other words, while observed
and unobserved pre-treatment variables such as literacy, education,
and ideology can gradually change, the assumption requires that they
should not discontinuously shift at the population cutoff.  The main advantage of
this design is internal validity, meaning that the resulting estimate
of causal effect at the threshold is credible.  The disadvantage,
however, is the lack of external validity, meaning that the estimate
we obtain may not be applicable beyond the region near the threshold.

## Question 2

Begin by creating a variable that determines how close each
  municipality was to the cutoff that determined whether states
  received a transfer or not. Transfers occurred at three separate
  population cutoffs: 10,188, 13,584, and 16,980. Using these cutoffs,
  create a single variable that characterizes the difference from the
  closest population cutoff.  Following the original analysis,
  standardize this measure by dividing the difference with the
  corresponding cutoff and multiply it by 100.  This will yield a
  normalized percent score for the difference between the population
  of each state and the cutoff relative to the cutoff value.

## Answer 2

We will use the midpoint between each of the cutoffs to determine
  which cutoff to use for a given observation.  If an observation is
  exactly at the midpoint, assign that observation the lower cutoff.

```{r}
## load data
data <- read.csv("data/transfer.csv")
mid1 <- 10188 + (13584 - 10188) / 2
mid2 <- 13584 + (16980 - 13584) / 2
## Create normalized percent score variable
data$pscore <-
    ifelse(data$pop82 <= mid1, (data$pop82 - 10188)/10188,
           ifelse(data$pop82 <= mid2, (data$pop82 - 13584)/13584,
                  (data$pop82 - 16980)/16980))*100
```

## Question 3

Begin by subsetting the data to include
  only those municipalities within 3 points of the funding cutoff on
  either side.  Using regressions, estimate the average causal effect
  of government transfer on each of the three outcome variables of
  interest: educational attainment, literacy, and poverty.  Give a
  brief substantive interpretation of the results.

## Answer 3

```{r}
dta.below <- subset(data, (data$pscore >= -3) & (data$pscore <= 0))
dta.above <- subset(data, (data$pscore >= 0) & (data$pscore <= 3))
## effect on literacy rate
lit.fit1 <- lm(literate91 ~ pscore, data = dta.below)
lit.fit2 <- lm(literate91 ~ pscore, data = dta.above)
coef(lit.fit2)[1] - coef(lit.fit1)[1]
## effect on educational attainment
educ.fit1 <- lm(educ91 ~ pscore, data = dta.below)
educ.fit2 <- lm(educ91 ~ pscore, data = dta.above)
coef(educ.fit2)[1] - coef(educ.fit1)[1]
## effect on poverty rate
pov.fit1 <- lm(poverty91 ~ pscore, data = dta.below)
pov.fit2 <- lm(poverty91 ~ pscore, data = dta.above)
coef(pov.fit2)[1] - coef(pov.fit1)[1]
```

Though the magnitude of effects is relatively small, increased
government transfers appear to improve educational attainment and
literacy while reducing the poverty rate.  The point estimates
obtained via regression can be interpreted as the average causal
effect of receiving a fiscal transfer at the population threshold. The
results suggest that the fiscal transfers increased the literacy rate
by `r round(coef(lit.fit2)[1] - coef(lit.fit1)[1], 2) * 100`
percentage points, increased the average educational attainment by
`r round(coef(educ.fit2)[1] - coef(educ.fit1)[1], 2)` years, and
reduced the poverty rate by `r -round(coef(pov.fit2)[1] -
  coef(pov.fit1)[1], 2)` percentage points.

## Question 4

Visualize the analysis done in the previous question by plotting
  data points, fitted regression lines, and the population threshold.
  Briefly comment on the plot.

## Answer 4

```{r}
par(cex = 1.5)
dta <- subset(data, (data$pscore >= -3) & (data$pscore <= 3))

## Literacy Rate Plot
plot(dta$pscore, dta$literate91,
     xlab = "Distance to Population Cutoff (%)",
     ylab = "Literacy Rate in 1991", main = "Literacy Rate")
abline(v = 0, col = "red", lty = "dashed")
lines(c(0, 3), predict(lit.fit2,
                       newdata = data.frame(pscore = c(0, 3))))
lines(c(-3, 0), predict(lit.fit1,
                        newdata = data.frame(pscore = c(-3, 0))))
```

```{r}
## Educational Attainment Plot
plot(dta$pscore, dta$educ91,
     xlab = "Distance to Population Cutoff (%)",
     ylab = "Avg Years Schooling in 1991",
     main = "Educational Attainment")
abline(v = 0, col = "red", lty = "dashed")
lines(c(0, 3), predict(educ.fit2,
                       newdata = data.frame(pscore = c(0, 3))))
lines(c(-3, 0), predict(educ.fit1,
                        newdata = data.frame(pscore = c(-3, 0))))
```

```{r}
## Poverty Rate Plot
plot(dta$pscore, dta$poverty91,
     xlab = "Distance to Population Cutoff (%)",
     ylab = "Poverty Rate in 1991", main = "Poverty Rate")
abline(v = 0, col = "red", lty = "dashed")
lines(c(0, 3), predict(pov.fit2,
                       newdata = data.frame(pscore = c(0, 3))))
lines(c(-3, 0), predict(pov.fit1,
                        newdata = data.frame(pscore = c(-3, 0))))
```

The figure visualizes the results we found in the previous
question. For literacy and educational attainment, we see that the
predicted value at the discontinuity using the regression model fitted
to the left side data is slightly lower than the predicted value using
the right side data. For the poverty rate measure, we see the opposite
is the case.

## Question 5
Instead of fitting linear regression models, we compute the
  difference in means of the outcome variables between the groups of
  observations above the threshold and below it.  How do the estimates
  differ from what you obtained in the earlier Question?  Is
  the assumption invoked here identical to the one required for the
  analysis conducted there?  Which
  estimates are more appropriate?  Discuss.

## Answer 5

```{r}
## diff-in-means
mean(dta.above$literate91) - mean(dta.below$literate91) # literacy rate
mean(dta.above$educ91) - mean(dta.below$educ91) # education
mean(dta.above$poverty91) - mean(dta.below$poverty91) # poverty
```

The difference-in-means estimates are based on an additional
assumption that the slopes of regressions below and above the
threshold are both zero (i.e., the regression lines are flat).  That
is, the assumption is that within the 3 percent window, the population
is not associated with the outcome variables.  If this assumption is
appropriate (is violated), these difference-in-means estimates are
similar to (different from) the regression-based estimates.  In
general, since the difference-in-means estimates require an additional
assumption, the regression-based estimates are more appropriate.

The results show that despite subtle differences, the two approaches
yield similar conclusions.  While the slopes of the regression lines
are not zero (especially for educational attainment and poverty rate),
the fact that the slopes have opposite signs and similar absolute
magnitude below and above the threshold means that the
difference-in-means estimators yield results similar to those based on
regressions.

## Question 6

Repeat the analysis conducted in the original question
  but vary the width of analysis window from 1 to 5 percentage points
  below and above the threshold.  Obtain the estimate for every
  percentage point.  Briefly comment on the results.

## Answer 6

```{r}
results <- matrix(NA, nrow = 5, ncol = 3)
colnames(results) <- c("literacy", "education", "poverty")
rownames(results) <- 1:5
## Replicate previous analysis for each window size
for(i in 1:5){ ## Loop over each of the 5 windows
  idta <- data[(data$pscore >= -i) & (data$pscore <= i), ]
  idta.below <- data[(data$pscore >= -i) & (data$pscore <= 0), ]
  idta.above <- data[(data$pscore >= 0) & (data$pscore <= i), ]
  ## Literacy rate
  lit.fit1 <- lm(literate91 ~ pscore, data = idta.below)
  lit.fit2 <- lm(literate91 ~ pscore, data = idta.above)
  lit.effect <- coef(lit.fit2)[1] - coef(lit.fit1)[1]
  ## Educational attainment
  educ.fit1 <- lm(educ91 ~ pscore, data = idta.below)
  educ.fit2 <- lm(educ91 ~ pscore, data = idta.above)
  educ.effect <- coef(educ.fit2)[1] - coef(educ.fit1)[1]
  ## Poverty rate
  pov.fit1 <- lm(poverty91 ~ pscore, data = idta.below)
  pov.fit2 <- lm(poverty91 ~ pscore, data = idta.above)
  pov.effect <- coef(pov.fit2)[1] - coef(pov.fit1)[1]
  ## Put into single vector and combine in the loop
  results[i, ] <- c(lit.effect, educ.effect, pov.effect)
}
results
```

The results change substantially as we vary the window we use to
perform the regression analysis. In general, the pattern is that as
the window increases, the estimated effects become larger. The largest
variation by window size occurs in the educational attainment
variable, while the literacy and poverty rate measures are more
stable. For all three estimated effects, however, the direction of
effects does not change as we vary the window size.

## Question 7

Conduct the same analysis as in the
  earlier Question but this time using measures of the
  poverty rate and educational attainment taken in 1980, before the
  population-based government transfers began. What do the results
  suggest about the validity of analysis presented in
  the earlier Question?
## Answer 7

```{r}
## Educational attainment in 1980
educ.fit1 <- lm(educ80 ~ pscore, data = dta.below)
educ.fit2 <- lm(educ80 ~ pscore, data = dta.above)
coef(educ.fit2)[1] - coef(educ.fit1)[1]
## Poverty rate in 1980
pov.fit1 <- lm(poverty80 ~ pscore, data = dta.below)
pov.fit2 <- lm(poverty80 ~ pscore, data = dta.above)
coef(pov.fit2)[1] - coef(pov.fit1)[1]
```

The use of measurements taken before the fiscal transfers began
amounts to a placebo test. If the RD design is valid, whether or not a
state received additional revenue by virtue of being above the
population cutoff should have no bearing on its poverty rate or
educational attainment *before those transfers began*.
Conducting the same analysis in Question 2, we see that the
difference in poverty rate is less than half the estimated difference
using the 1991 measure. Similarly, the average years of school is
approximately a third as large as what we estimate using the 1991
measure.  These differences are small relative to what we observe with
the 1991 measure and thus suggest that the discontinuity design is
valid. To be more certain, however, we would need to determine whether
the estimates are statistically distinguishable from zero, a topic
covered in later chapters.

