---
title: "Long-Term Effects of 9/11 on the Political Behavior of Victims' Families"
output:
  pdf_document: default
  html_document: default
---

In this exercise, we examine a hypothesis that individuals who lost
someone in the terrorist attacks of 9/11, whether a family relative or
a neighbor, will have become more politically engaged.

This exercise is based on: Hersh, E. D. 2013. “[Long-Term Effect of September 11 on the Political Behavior of Victims’ Families and Neighbors.](http://dx.doi.org/10.1073/pnas.1315043110)” *Proceedings of the National Academy of Sciences* 110(52): 20959–63. 

We will examine this hypothesis
using several different estimation techniques, focusing throughout on
the effect of the attacks on the victims' families rather than their
neighbors. The CSV data file, `victims9-11.csv`, contains the
following variables:

------------------------------------------------------------------------------
 Name                Description
 ------------------- ---------------------------------------------------------
 `voter.id`          Unique identifiers of relatives and neighbors of the 
                     victims 
                     
 `treatment`         Families and neighbors of actual victims (1) vs control 
                     group (0)
 
 `victim.status`     Families (2) vs neighbors (3) of victims and controls
 
 `ge20xx`            Voting in the `20xx` general election (`Y`=at the polls, 
                     `A`=absentee, `E`=early, `M`=by mail) 
 
 `fam.members`       Number of family members living with voter at their address 
 
 `age`               Voter's age 
 
 `party`             Voter's party affiliation (`D`=Democrat, `R`=Republicans, 
                     `N`=no affiliation) 
 
 `sex`               Voter's sex
 
 `pct.white`         Proportion of non-Hispanic white voters living on the same 
                     block 
 
 `median.income`     Median income of voters living on the same block  
------------------------------------------------------------------------------

Voters were
included in the data on the basis of their relationship to actual
victims - these constitute the two treatment groups - or if no
such relationship existed but they were, otherwise, sufficiently
similar to voters in the treatment groups - this constitutes the
control group. 

## Question 1

We begin by reformatting the data to facilitate our analysis.
The three variables that contain non-numerical values are sex,
party, and voting records. Rewrite these variables using the
following coding rules: for the party variable, assign the value of
$1$ to all Democrats, $-1$ to all Republicans, and $0$ for all other
parties and non-affiliated citizens; for the sex variable, assign
the value of $1$ to all female citizens and $0$ to all male
citizens; finally, for each of the seven voting record variables,
assign the value of $1$ if the individual voted in a given election,
and $0$ otherwise.

## Answer 1
```{r}
victims <- read.csv("data/911-victims.csv")

# Recode the party and sex variables
victims$party <- ifelse(victims$party == "D", 1, 
                        ifelse(victims$party == "R", -1, 0))
victims$sex <- ifelse(victims$sex == "F", 1, 0)

# Create a vector denoting the unique general election years
years <- seq(from= 2000, to=2012, by=2)

## Recode each of the seven turnout variables,
## where the loop isolates one turnout variable at a time
for (i in 1:length(years)) {
  voting.rec <- victims[,paste("ge", years[i], sep="")]
  victims[,paste("ge", years[i], sep="")] <- 
    ifelse(voting.rec == "A" | voting.rec == "Y" | 
             voting.rec == "E" | voting.rec == "M", 1, 0)
}
```

## Question 2

We will now test the hypothesis. Calculate the sample Average
Treatment Effect of the terrorist attacks on voter turnout
separately for each of the seven general elections between 2000 and
2012. Focus on families of victims as the treatment, and disregard
the neighbor category. For each of the seven point-estimates,
compute the corresponding standard error.  Assume that both the
treatment and control groups form two random samples, and that the
two samples are statistically independent. Then, using the quantiles
of the standard normal distribution, calculate the 95% confidence
intervals for each of the seven point-estimates. Plot the results
where the horizontal axis represents the elections.  Provide a brief
interpretation, with a particular focus on the meaning of the
confidence intervals you computed.

## Answer 2

```{r}
diff.in.means <- vector()
std.errors <- vector()

for (i in 1:length(years)) {
  voting.rec <- victims[,paste("ge", years[i], sep="")]
  treatment <- na.omit(voting.rec[victims$treatment == 1 & 
                                 victims$victim.status == 2])
  control <- na.omit(voting.rec[victims$treatment == 0 & 
                                 victims$victim.status == 2])
  
  diff.in.means[i] <- mean(treatment) - mean(control)
  
  std.errors[i] <- sqrt(var(treatment)/length(treatment) +
                       var(control)/length(control))
}
par(mar=c(4,5,3,2))
plot(years, diff.in.means, pch = 17, col = "red",
     ylim = c(-0.04, 0.08), ylab = "Difference in Turnout,
     Family - Control", xlab = "Year of General Election",
     main = "Victims' Families Relative to Control")

abline(h=0)
abline(v=2001.75)
text("Attacks of September 11", x=2004.4, y=0.06)

segments(x0=years, y0=diff.in.means - std.errors * 1.96,
         x1=years, y1=diff.in.means + std.errors * 1.96, 
         lty=2)
```

The ATE for the year 2000, before the attacks of 9/11, functions as 
a placebo for the validity of our estimation strategy. If the treatment
and control groups were balanced prior to the administration of treatment,
then the ATE for 2000 should yield an effect of exactly 0, which is what we
observe. 

Furthermore, with the exception of the 2004 election, we observe a 
consistently positive effect of the attacks on voter turnout, and it
does not appear to decay with time. Subject to the validity of ATE
as estimation strategy (that is, subject to the assumption that the
treatment and control groups were balanced prior to the administration
of the treatment), this analysis lends support to the acitivation hypothesis.

Nevertheless, we notice that the confidence intervals for the 2002, 2004,
2006, and 2008 elections include the value of 0 (null effect). This means
that our causal effect estimates for those years are statistically 
indistinguishable from 0 at the 5% significance level. The only
significant results are for the years of 2010 and 2012, nearly ten 
years after the attacks.

## Question 3

To examine the validity of the cross-sectional comparisons
conducted in Question 2, check whether possible confounders are
balanced between the treatment and control groups. Compare the means
of the last six variables in the table above across the two
groups along with their 95\% confidence intervals.  Provide a brief
interpretation of the results. What can you conclude about the
validity of the cross-sectional comparisons?

## Answer 3

```{r}
covariates <- colnames(victims)[11:16]
cov.balance <- vector()
cov.se <- vector()

for (i in 1:length(covariates)) {
  covariate <- victims[,covariates[i]]
  treatment <- na.omit(covariate[victims$treatment == 1 & 
                                 victims$victim.status == 2])
  control <- na.omit(covariate[victims$treatment == 0 & 
                                 victims$victim.status == 2])
  
  cov.se[i] <- sqrt(var(treatment) / length(treatment) + 
                   var(control) / length(control))
  
  cov.balance[i] <- mean(treatment) - mean(control)
}

## Calculate the difference in means
cov.balance

## Calculate the lower bound of the 
## confidence intervals
cov.balance - qnorm(0.975) * cov.se

## Calculate the upper bound of the 
## confidence intervals
cov.balance + qnorm(0.975) * cov.se
```

If the treatment and control groups are properly randomized,
the means on the six possible confounders should be similar
across the two groups. While none of the point-estimates are
exactly zero, the value of zero is included in each of the six
confidence intervals. That means that, at the 5% significance
level, the means of the two groups are statistically 
indistinguishable from one another across the six variables. 
The magnitude of observed differences between them is consistent
with what we would expect to see purely by chance. 

This confirms the validity of the ATE as an unbiased estimate of
the true causal effect, as calculated in the previous question.

## Question 4

Now, focus on the treatment group only. Compute the
before-and-after estimate of the effect of the terrorist attacks on
voter turnout for families of the victims by using the 2000 general
election as a baseline for each of the subsequent six general
elections. Next, calculate the standard errors corresponding to the
six point-estimates.  Using the standard errors, compute the 95%
confidence intervals for each of the six point-estimates.  Plot the
results. Provide a brief interpretation of your
findings. **Hint**: Since you are now tracing the
same sample evolve through time, you can no longer assume the
independence of sample means in calculating the standard errors;
take care to incorporate the covariance between voter turnout in 2000
and a subsequent election into your calculations.

## Answer 4
```{r}
before.after <- vector()
std.errors.ba <- vector()

for (i in 2:length(years)) {
  voting.rec <- na.omit(victims[victims$treatment == 1 & 
                                 victims$victim.status == 2
                        ,c("ge2000", 
                           paste("ge", years[i], sep=""))])
  
  treatment.before <- voting.rec[,1]
  treatment.after <- voting.rec[,2]
  
  before.after[i-1] <- mean(treatment.after - treatment.before)
  
  n <- length(treatment.before)
  
  std.errors.ba[i-1] <- sqrt(var(treatment.after)/n +
                       var(treatment.before)/n - 
                         (2/n) * cov(treatment.after, treatment.before))
}

par(mar=c(4,5,3,2))
plot(years[-1], before.after, pch = 17, col = "red",
     ylim = c(-0.25, 0.08), ylab = "Difference in Turnout,
     After - Before (2000)", xlab = "Year of General Election",
     main = "Longitudinal Analysis of Victims' Families\n (treatment only)")

abline(h=0)

for (i in 2:length(years)) {
  segments(x0=years[i], y0=before.after[i-1] - 
             std.errors.ba[i-1] * 1.96,
           x1=years[i], y1=before.after[i-1] + 
             std.errors.ba[i-1] * 1.96,
           lty=2)
}
```

We can see that, with the exception of the estimate for the 2012
election, all estimates are statistically significant (that is,
statistically distinguishable from 0) at the 5% level. An 
interesting trend emerges: following the 9/11 attacks, using
the turnout in the 2000 election as our baseline, turnout
increases for the subsequent presidential election years, with
the exception of 2012, and declines for the midterm election
years.

Notice that we should be wary of endowing the before-and-after
estimates with causal interptretations, as this overlooks the 
possibility that the observed temporal trend may be consistent 
with the one observed for the control group.

## Question 5

We will now repeat the same analysis as in Question 4, but now
using voters in the control group only. As before, be sure to
restrict your analysis to family relatives of control victims rather
than their neighbors.  What does this analysis imply about the
validity of the analysis in the previous question?

## Answer 5

```{r}
before.after.ctrl <- vector()
std.errors.ba.ctrl <- vector()

for (i in 2:length(years)) {
  voting.rec <- na.omit(victims[victims$treatment == 0 & 
                                 victims$victim.status == 2,
                        c("ge2000", paste("ge", years[i], sep=""))])
  
  ctrl.before <- voting.rec[,1]
  ctrl.after <- voting.rec[,2]
  
  before.after.ctrl[i-1] <- mean(ctrl.after - ctrl.before)
  
  n <- length(ctrl.before)
  
  std.errors.ba.ctrl[i-1] <- sqrt(var(ctrl.after)/n +
                       var(ctrl.before)/n - 
                         2/n * cov(ctrl.after, ctrl.before))
}

par(mar=c(4,5,3,2))
plot(years[-1], before.after.ctrl, pch = 17, col = "red",
     ylim = c(-0.25, 0.05), ylab = "Difference in Turnout,
     After - Before", xlab = "Year of General Election",
     main = "Longitudinal Analysis of Victims' Families\n (control only)")

abline(h=0)

for (i in 2:length(years)) {
  segments(x0=years[i], y0=before.after.ctrl[i-1] - std.errors.ba.ctrl[i-1] * 1.96,
           x1=years[i], y1=before.after.ctrl[i-1] + std.errors.ba.ctrl[i-1] * 1.96,
           lty=2)
}
```

With the exception of the 2008 election, none of the confidence
intervals contain the value of 0, so all of the other five
estimates are statistically significant at the 5% level. Notice
that the temporal trend of the before-and-after estimates are similar
for the treatment and control groups, which strengthens our
suspicion of the validity of the before-and-ater comparison for 
the treatment group. 

To isolate the effect of the treatment in a
temporal perspective, we need to pool information from both 
the treatment and control groups and account for the similarity
of their temporal paths. Difference-in-differences is a suitable 
estimation strategy for that.

## Question 6

Calculate the difference-in-differences estimate for the general
elections immediately surrounding the attacks, 2000 and
2002. Calculate the standard error for your estimate, and provide
the 95% confidence interval.  Once again, you are allowed to use
the quantiles of the standard normal distribution in calculating the
confidence interval.  Provide a brief interpretation of the result.

## Answer 6

```{r}
DiD <- before.after[1] - before.after.ctrl[1]
DiD

DiD.se <- sqrt(std.errors.ba[1]^2 + 
                 std.errors.ba.ctrl[1]^2)

ci <- c(DiD - DiD.se * 1.96, DiD + DiD.se * 1.96)
ci
```

## Question 7

Repeat the previous difference-in-differences analysis for all
six elections following the attacks.  Throughout this analysis,
use the 2000 election as the baseline as done in the previous
question.  Plot the results with the horizontal axis representing
different elections.  Provide a substantive interpretation of the
results.

## Answer 7
```{r}
DiDs <- before.after - before.after.ctrl
DiDs

DiDs.se <- sqrt(std.errors.ba^2 + std.errors.ba.ctrl^2)
DiDs.se

par(mar=c(4,5,3,2))
plot(years[-1], DiDs, pch = 17, col = "red",
     ylim = c(-0.04, 0.08), ylab = "Difference in Turnout", 
     xlab = "Year of General Election",
     main = "Longitudinal Analysis of Victims' Families
     (Difference in Differences)")

abline(h=0)

for (i in 2:length(years)) {
  segments(x0=years[i], y0=DiDs[i-1] - DiDs.se[i-1] * 1.96,
           x1=years[i], y1=DiDs[i-1] + DiDs.se[i-1] * 1.96,
           lty=2)
}
```

The DiD estimates are similar to the SATE estimates, which is precisely
what we would expect given the results of our placebo test in Question
2. With the exception of the 2004 election, we observe a consistently
positive effect, although the effect is only statistically significant at
the 5% level for the 2010 and 2012 elections. 
