---
title: "Efficacy of Small-class Size in Early Education"
output:
  html_document: default
  pdf_document: default
date: "5 August 2015"
---

The STAR (Student-Teacher Achievement Ratio) Project is a four year
*longitudinal study* examining the effect of class size in early
grade levels on educational performance and personal
development. 

This exercise is in part based on:
 Mosteller, Frederick. 1997. “[The Tennessee Study of Class Size in the 
 Early School Grades.](http://dx.doi.org/10.2307/3824562)” *Bulletin of 
 the American Academy of Arts and Sciences* 50(7): 14-25.
  
A longitudinal study is one in which the same
participants are followed over time.  This particular study lasted
from 1985 to 1989 involved 11,601 students. During the four years of
the study, students were randomly assigned to small classes,
regular-sized classes, or regular-sized classes with an aid.  In all,
the experiment cost around $12 million. Even though the program
stopped in 1989 after the first kindergarten class in the program
finished third grade, collection of various measurements (e.g.,
performance on tests in eighth grade, overall high school GPA)
continued through the end of participants' high school attendance.

We will analyze just a portion of this data to investigate whether the
small class sizes improved performance or not. The data file name is
`STAR.csv`, which is a CSV data file.  The names and
descriptions of variables in this data set are:

--------------------------------------------------------------------------------
 Name                 Description
 -------------------- ----------------------------------------------------------
 `race`               Student's race (White = 1, Black = 2, Asian = 3, 
                      Hispanic = 4,  Native American = 5, Others = 6)
 
 `classtype`          Type of kindergarten class (small = 1, regular = 2, regular with aid = 3)
 
 `g4math`             Total scaled score for math portion of fourth grade standardized test 
 
 `g4reading`          Total scaled score for reading portion of fourth grade standardized test 
 
 `yearssmall`         Number of years in small classes 
 
 `hsgrad`             High school graduation (did graduate = 1, did not graduate = 0) 
 --------------------------------------------------------------------------------
 
Note that there are a fair amount of missing
values in this data set.  For example, missing values arise because
some students left a STAR school before third grade or did not enter a
STAR school until first grade.

## Question 1

Create a new factor variable called `kinder` in the data
  frame.  This variable should recode `classtype` by changing
  integer values to their corresponding informative labels (e.g.,
  change 1 to `small` etc.).  Similarly, recode the
  `race` variable into a factor variable with four levels
  (`white`, `black`, `hispanic`, `others`) by
  combining Asians and Native Americans as the `others`
  category.  For the `race` variable, overwrite the original
  variable in the data frame rather than creating a new one.  Recall
  that `na.rm = TRUE` can be added to functions in order to
  remove missing data.
  
## Answer 1

```{r}
## read in data
STAR <- read.csv("data/STAR.csv")
## create a variable labeling class type
STAR$kinder <- ifelse(STAR$classtype == 1, "small", 
                      ifelse(STAR$classtype == 2, "regular", 
                             "reg w/aid"))
STAR$kinder <- as.factor(STAR$kinder)

## create race label variable
race <- ifelse(STAR$race == 1, "white",
               ifelse(STAR$race == 2, "black",
                      ifelse(STAR$race == 4, "hispanic", "others")))
race <- as.factor(race)
## overwrite the original variable
STAR$race <- race
```

## Question 2

How does performance on fourth grade reading and math tests for
  those students assigned to a small class in kindergarten compare
  with those assigned to a regular-sized class?  Do students in the
  smaller classes perform better?  Use means to make this comparison
  while removing missing values.  Give a brief substantive
  interpretation of the results.  To understand the size of the
  estimated effects, compare them with the standard deviation of the
  test scores.
   
## Answer 2

```{r}
## diff in means for reading
mean(STAR$g4reading[STAR$kinder == "small"], na.rm = TRUE) - 
    mean(STAR$g4reading[STAR$kinder == "regular"], na.rm = TRUE)
## diff in means for math
mean(STAR$g4math[STAR$kinder == "small"], na.rm = TRUE) - 
    mean(STAR$g4math[STAR$kinder == "regular"], na.rm = TRUE)
```

This output shows that students in small classrooms in kindergarten
scored, on average, `r round(mean(STAR$g4reading[STAR$kinder ==
  "small"], na.rm = TRUE) - mean(STAR$g4reading[STAR$kinder ==
  "regular"], na.rm = TRUE), 2)` points higher on reading tests in the
fourth grade.  However, they scored about
`r round(mean(STAR$g4math[STAR$kinder == "small"], na.rm = TRUE) -
  mean(STAR$g4math[STAR$kinder == "regular"], na.rm = TRUE), 2)*(-1)`
points lower on average in math compared to students in regular
classrooms.  Comparing these estimated effects with the standard
deviation of test scores, we observe that the sizes of the estimated
effects are quite small.

```{r}
## standard deviations of reading and math scores
sd(STAR$g4reading, na.rm = TRUE)
sd(STAR$g4math, na.rm = TRUE)
```

We conclude that on average small class size in kindergarden did not
substantially increase reading and math test scores in the fourth
grade.

## Question 3

Instead of comparing just average scores of reading and math
  tests between those students assigned to small classes and those
  assigned to regular-sized classes, look at the entire range of
  possible scores.  To do so, compare a high score, defined as the
  66th percentile, and a low score (the 33rd percentile) for small
  classes with the corresponding score for regular classes.  These are
  examples of *quantile treatment effects*.  Does this analysis
  add anything to the analysis based on mean in the previous question?

## Answer 3

```{r}
## reading scores
reading.small <- quantile(STAR$g4reading[STAR$kinder == "small"], 
                          na.rm = TRUE, probs = c(1/3, 2/3) )
reading.regular <- quantile(STAR$g4reading[STAR$kinder == "regular"], 
                          na.rm = TRUE, probs = c(1/3, 2/3))
reading.small - reading.regular  # difference in quantiles
## math scores
math.small <- quantile(STAR$g4math[STAR$kinder == "small"], 
                       na.rm = TRUE, probs = c(0.33, 0.66, 1))
math.regular <- quantile(STAR$g4math[STAR$kinder == "regular"], 
                          na.rm = TRUE, probs = c(0.33, 0.66, 1))
math.small - math.regular
```

The differences in reading and math scores between students in small
classes and regular classes were very small or nonexistent at the 33rd
and 66th percentiles.  This analysis therefore confirms the conclusion
of the analysis in the previous question that small class size in
kindergarten did little to increase test scores at the fourth grade
level.

## Question 4

Some students were in small classes for all four years that the
  STAR program ran. Others were assigned to small classes for only one
  year and had either regular classes or regular classes with an aid
  for the rest. How many such students of each type are in the data
  set?  Create a contingency table of proportions using the
  `kinder` and `yearsmall` variables.  Does participation
  in more years of small classes make a greater difference in test
  scores?  Compare the average and median reading and math test scores
  across students who spent different numbers of years in small
  classes.

## Answer 4

```{r}
## contingency table of proportions
prop.table(table(STAR$kinder, STAR$yearssmall))
## mean/median reading score across years in small classes
tapply(STAR$g4reading, STAR$yearssmall, mean, na.rm = TRUE) 
tapply(STAR$g4reading, STAR$yearssmall, median, na.rm = TRUE) 
## mean/median math score
tapply(STAR$g4math, STAR$yearssmall, mean, na.rm = TRUE) 
tapply(STAR$g4math, STAR$yearssmall, median, na.rm = TRUE) 
```

The contingency table shows that
`r round(sum(prop.table(table(STAR$kinder, STAR$yearssmall))[,
  1]), 2)*100`% of students were never in the small classes whereas
`r round(prop.table(table(STAR$kinder, STAR$yearssmall))["small",
  5], 2)*100`% of them were in the small classes for all four years.
The analysis suggests that in general, spending all four years in the
small classes increases both reading and math test scores.  The effect
is not huge but is of reasonable size.  The analyses based on the mean
and median yield similar results.

## Question 5

We examine whether the STAR program reduced the achievement gaps
  across different racial groups.  Begin by comparing the average
  reading and math test scores between white and minority students
  (i.e., Blacks and hispanics) among those students who were assigned
  to regular classes with no aid.  Conduct the same comparison among
  those students who were assigned to small classes.  Give a brief
  substantive interpretation of the results of your analysis.
  
## Answer 5

```{r}
## subset the data
white <- subset(STAR, subset = (race == "white"))
minority <- subset(STAR, subset = ((race == "black") | 
                                       (race == "hispanic")))
## racial gap for reading among students in regular classes
mean(white$g4reading[white$kinder == "regular"], na.rm = TRUE) - 
         mean(minority$g4reading[minority$kinder == "regular"], 
              na.rm = TRUE)
## racial gap for reading among students in small classes
mean(white$g4reading[white$kinder == "small"], na.rm = TRUE) - 
    mean(minority$g4reading[minority$kinder == "small"], 
              na.rm = TRUE)
## repeat the same analysis for math score
mean(white$g4math[white$kinder == "regular"], na.rm = TRUE) - 
    mean(minority$g4math[minority$kinder == "regular"], 
         na.rm = TRUE)
mean(white$g4math[white$kinder == "small"], na.rm = TRUE) - 
    mean(minority$g4math[minority$kinder == "small"], 
         na.rm = TRUE)
```

Our analysis shows that there is a substantial racial gap.  On
average, white students tend to perform better in both reading and
math scores than minority students, regardless of class sizes.
However, in terms of reading test scores, this achievement gap is
reduced when students are assigned to small classes.  This suggests
that minority students benefited more from small classes than white
students.  The same conclusion, however, does not apply to math
scores.  The racial achievement gap is approximately the same size in
both regular and small classes.

## Question 6

We consider the long term effects of kindergarden class size.
  Compare high school graduation rates across students assigned to
  different class types.  Also, examine whether graduation rates
  differ by the number of years spent in small classses.  Finally, as
  done in the previous question, investigate whether the STAR program
  has reduced the racial gap between white and minority students'
  graduation rates.  Briefly discuss the results.

## Answer 6

```{r}
## graduation rate by classsize
tapply(STAR$hsgrad, STAR$kinder, mean, na.rm = TRUE)
tapply(STAR$hsgrad, STAR$yearssmall, mean, na.rm = TRUE)
```

We observe little difference in graduation rates across students who
were assigned to different class types in kindergarden.  However,
those who spent all four years in small classes have a higher
graduation rate than the others.  This result is consistent with the
analysis for a previous question where we found spending all four
years in small classes increases the average reading score.

```{r}
## racial gap 
tapply(white$hsgrad, white$kinder, mean, na.rm = TRUE) -
    tapply(minority$hsgrad, minority$kinder, mean, na.rm = TRUE)
tapply(white$hsgrad, white$yearssmall, mean, na.rm = TRUE) -
    tapply(minority$hsgrad, minority$yearssmall, mean, na.rm = TRUE)
```

We observe substantial racial gaps in high school graduation rates
regardless of kindergraden class types.  These gaps appear to remain
even when students were assigned to small classes for all four years.
Therefore, the STAR program appears to have little impact in closing
racial gaps of high school graduation rates.


