---
title: "Prediction Based on Betting Markets"
output: pdf_document
---

Earlier in the chapter, we studied the prediction of election outcomes
using polls.  Here, we study the prediction of election outcomes based
on betting markets.  In particular, we analyze data for the 2008 and
2012 US presidential elections from the online betting company, called
Intrade.  At Intrade, people trade contracts such as 'Obama to win
the electoral votes of Florida.'  Each contract's market price
fluctuates based on its sales.  Why might we expect betting markets
like Intrade to accurately predict the outcomes of elections or of
other events?  Some argue that the market can aggregate available
information efficiently.  In this exercise, we will test this 
*efficient market hypothesis* by analyzing the market prices of
contracts for Democratic and Republican nominees' victories in each
state.

The data files for 2008 and 2012 are available in CSV format as
`intrade08.csv` and `intrade12.csv`, respectively. The variables in these datasets are:

--------------------------------------------------------------------------------
 Name                Description
 ------------------- ----------------------------------------------------------
 `day`               Date of the session 
 
 `statename`         Full name of each state (including District of 
                     Columbia in 2008) 
 
 `state`             Abbreviation of each state (including District of 
                     Columbia in 2008)
 
 `PriceD`            Closing price (predicted vote share) of Democratic 
                     Nominee's market
 
 `PriceR`            Closing price (predicted vote share) of Republican 
                     Nominee's market 
 
 `VolumeD`           Total session trades of Democratic Party Nominee's 
                     market
 
 `VolumeR`           Total session trades of Republican Party Nominee's 
                    market 
--------------------------------------------------------------------------------

Each row represents daily trading
information about the contracts for either the Democratic or
Republican Party nominee's victory in a particular state. 

We will also
use the election outcome data.  These data files are
`pres08.csv` and `pres12.csv` with variables:

--------------------------------------------------------------------------------
 Name                 Description
 -------------------- ----------------------------------------------------------
 `state.name`         Full name of state (only in `pres2008`)
 
 `state`              Two letter state abbreviation
 
 `Obama`              Vote percentage for Obama
 
 `McCain`             Vote percentage for McCain
 
 `EV`                 Number of electoral college votes for this state
--------------------------------------------------------------------------------
 
We'll also use poll data from 2008 and 2012 in the files `polls08.csv` and `polls12.csv`,  The variables in the polling data are:
  
--------------------------------------------------------------------------------
 Name                 Description
 -------------------- ---------------------------------------------------------
 `state`              Abbreviated name of state in which poll was conducted 
 
 `Obama`              Predicted support for Obama (percentage)
 
 `Romney`             Predicted support for Romney (percentage) 
 
 `Pollster`           Name of organization conducting poll 
 
 `middate`            Middle of the period when poll was conducted 
--------------------------------------------------------------------------------

## Question 1

We will begin by using the market prices on the day before the
  election to predict the 2008 election outcome.  To do this, subset
  the data such that it contains the market information for each state
  and candidate only on the day before the election.  Note that in
  2008 the election day was November 4.  We compare the closing prices
  for the two candidates in a given state and classify a candidate
  whose contract has a higher price as the predicted winner of that
  state.  Which states were misclassified?  How does this compare to
  the classification by polls presented earlier in this chapter?
  Repeat the same analysis for the 2012 election, which was held on
  November 6.  How well did the prediction market do in 2012 compared
  to 2008?  Note that in 2012 some less competitive states have
  missing data on the day before the election because there were no
  trades on the Republican and Democratic betting markets. Assume
  Intrade predictions would have been accurate for these states.

## Answer 1

We begin by analyzing the 2008 data.

```{r}
intrade08 <- read.csv("data/intrade08.csv")
pres08 <- read.csv("data/pres08.csv")
intresults08 <- merge(intrade08, pres08, by = "state")

head(intresults08) # check the merged data
## Create Days To Election Variable for 2008
intresults08$DaysToElection <- as.Date("2008-11-04") - 
    as.Date(intresults08$day)
## Find difference in market prices to identify candidate 
## with higher price
intresults08$obama.intmarg <- intresults08$PriceD - 
                                      intresults08$PriceR
## Calculate actual result
intresults08$obama.actmarg <- intresults08$Obama - 
    intresults08$McCain
## Subset last day before election
latest08 <- intresults08[intresults08$DaysToElection == 1,]
## Identify states classified and misclassified
latest08$state[sign(latest08$obama.intmarg) != 
                   sign(latest08$obama.actmarg)]
```

We repeat the same analysis for the 2012 data.  Note that there are
some missing data

```{r}
intrade12 <- read.csv("data/intrade12.csv")
pres12 <- read.csv("data/pres12.csv")
intresults12 <- merge(intrade12, pres12, by = "state")

## Create Days To Election Variable for 2012
intresults12$DaysToElection <- as.Date("2012-11-06") - 
    as.Date(intresults12$day)
## Find difference in market prices to identify candidate 
## with higher price
intresults12$obama.intmarg <- intresults12$PriceD - 
    intresults12$PriceR
## Calculate actual result
intresults12$obama.actmarg <- intresults12$Obama - 
    intresults12$Romney
## Subset last day before election
latest12 <- intresults12[intresults12$DaysToElection == 1,]
## Identify states classified and misclassified
## missing data appear as `NA'
latest12$state[sign(latest12$obama.intmarg) != 
                   sign(latest12$obama.actmarg)]
## Number of states with missing data in 2012
sum(is.na(latest12$obama.intmarg))
```

In 2008, two states - Indiana and Missouri - out of 51 (including
DC) were misclassified by the prediction markets. This outperforms
the poll predictions from the chapter in which 3 out of 51 states were
misclassified. In 2012, only Florida was misclassified by the
Intrade market. There were 10 states with missing data which we assume
would have been accurately classified.

## Question 2

How do the predictions based on the betting markets change over
  time?  Implement the same classification procedure as above on each
  of the last 90 days of the 2008 campaign rather than just the day
  before the election.  Plot the predicted number of electoral votes
  for the Democratic party nominee over this 90-day period.  The
  resulting plot should also indicate the actual election result.
  Note that in 2008, Obama won 365 electoral votes. Briefly comment on
  the plot.

## Answer 2

```{r}
par(cex = 1.25)
## Empty vector to store predictions
Obama.dailypredEV08 <- rep(NA, 90)
for (i in 1:90) {
    ## Subset by i days to the election    
    daily.data  <- subset(intresults08, 
                          subset =  (DaysToElection == i))
    ## Sum Obama's EV for states with positive price margins   
    Obama.dailypredEV08[i] <- 
        sum(daily.data$EV[daily.data$obama.intmarg > 0])
} 
## Plot Predicted Electoral Votes from Intrade 
plot(1:90, Obama.dailypredEV08,
     main = "Daily Predicted Obama Electoral Votes",
     xlab = "Days to Election", ylab = "Predicted Electoral Votes",
     xlim = c(90, -1), ylim = c(260, 380))
points(0, 365, pch = 19)
abline(v = 0)
abline(h = 365, lty = "dashed")
text(80, 370, "Actual Result")  
```
  
Intrade predictions became more accurate as the election neared. While
the predictions were under-estimating Obama's total number of
electoral votes by 50 -- 100 votes with two to three months left in
the campaign, the prediction markets came within 10 electoral votes of
the actual total 20 days before the election.

## Question 3

Repeat the previous exercise but this time use the seven-day
  *moving-average* price, instead of the daily price, for each
  candidate within a state.  This can be done with a
  loop. For a given day, we take the average of the Session Close
  prices within the past seven days (including that day).  To answer
  this question, we must first compute the seven-day average within
  each state. Next, we sum the electoral votes for the states Obama is
  predicted to win.  Using the `tapply` function will allow us
  to efficiently compute the predicted winner for each state on a
  given day.
  
## Answer 3

```{r}
par(cex = 1.25)
## Empty vectors to store predictions
Obama.predEV <- rep(NA, 90)
for (i in 1:90) {
  ## Subset 7 days around day i    
  week.data  <- subset(intresults08, 
                       subset =  (DaysToElection <= (90 - i + 7)) & 
                           (DaysToElection > (90 - i)))
  ## Take average price margin within each state
  ## Bind this to the number of EV for each state
  week.means <- cbind(tapply(week.data$obama.intmarg, 
                             week.data$state, mean, na.rm = TRUE),
                      tapply(week.data$EV, week.data$state, 
                             mean, na.rm = TRUE))
  ## Subset only states where Obama is predicted to win
  week.obwin <- week.means[week.means[, 1] > 0, ]
  ## Sum Obama's predicted EV on a given day     
  Obama.predEV[i] <- sum(week.obwin[, 2])
} 
## Plot Predicted Electoral Votes from intrade 
plot(90:1, Obama.predEV,
     main = "7-Day Average Predicted Obama Electoral Votes",
     xlab = "Days to Election", ylab = "Predicted Electoral Votes",
     xlim = c(90, -1), ylim = c(260, 380))
points(0, 365, pch = 19)
abline(v = 0)
abline(h = 365, lty = "dashed")
text(80, 368, "Actual Result")
```

Like the previous exercise, the predictions underestimate Obama's
total number of electoral votes earlier in the campaign. However, 30
days before the election, the electoral vote predictions become very
accurate -- within a single vote of the actual total. These results
are more stable than using the single-day predictions from the
previous exercise.

## Question 4

Create a similar plot for 2008 state-wide poll predictions using
  the data file `polls08.csv`. Notice that polls are not conducted daily
  within each state. Therefore, within a given state for each of the
  last 90 days of the campaign, we compute the average margin of
  victory from the most recent poll(s) conducted. If multiple polls
  occurred on the same day, average these polls. Based on the most
  recent predictions in each state, sum Obama's total number of
  predicted electoral votes. One strategy to answer this question is
  to program two loops - an inner loop with 51 iterations for each
  state and an outer loop with 90 iterations for each day.
  
## Answer 4

```{r}
par(cex = 1.25)
## Load Poll data
polls08 <- read.csv("data/polls08.csv")
pollresults08 <- merge(polls08, pres08, by = "state")
## Calculate margins variables. 
## Note the suffixes .x and .y from the merge
pollresults08$obama.pollmarg <-pollresults08$Obama.x - 
    pollresults08$McCain.x
## Create DaysToElection variable
pollresults08$DaysToElection <- as.Date("2008-11-04") - 
    as.Date(pollresults08$middate)
## Empty vectors to store predicted EV
totpollEV <- rep(NA, 90)    
st.names <- unique(pollresults08$state) # To subset by state
## Iterate through each of 90 j days
for(j in 1:90){
  ## Use this object to compute total EV Obama is predicted to win 
  ## Counter resets to 0 after each iteration so that the total 
  ## number of Obama's predicted EV is calculated for each day
  counter <- 0
  ## Subset only observations j or more days from the election    
  near.data  <- subset(pollresults08, 
                       subset =  (DaysToElection >= j))
   ## Iterate through each of 51 i states
   for(i in 1:51) {
       ## Subset within state 
       state.data <- subset(near.data, subset = (state == st.names[i]))
       ## Subset most recent polls
       state.data <- 
           subset(state.data, 
                  subset = (DaysToElection == 
                                min(state.data$DaysToElection)))
       ## If Obama is predicted to win, add EV to the counter   
       if(mean(state.data$obama.pollmarg) > 0) {
           ## Overwrite the counter to update total after each state   
           counter <- counter + mean(state.data$EV)
       } 
   }
  ## Store predicted electoral votes from each day
  totpollEV[j] <- counter   
}
## Plot Predicted Electoral Votes from Polls
plot(1:90, totpollEV,
     main = "Poll Predicted Obama Electoral Votes",
     xlab = "Days to Election", ylab = "Predicted Electoral Votes",
     xlim = c(90, -1), ylim = c(260, 380))
points(0, 365, pch = 19)
abline(v = 0)
abline(h = 365, lty = "dashed")
text(80, 368, "Actual Result")
```

Predictions appear more erratic in polls than in prediction markets.
However, within the last month of the campaign, polls consistently
predicted Obama would win.  The week prior to the election, the polls
predicted outcomes within 16 votes of Obama's actual total number of
electoral votes in the election.  Nevertheless, over the same time
period, the prediction markets outperform polls, predicting within one
electoral vote of the actual total. 

## Question 5

What is the relationship between the price margins of the
  Intrade market and the actual margin of victory? Using only the
  market data from the day before the election in 2008, regress
  Obama's actual margin of victory in each state on Obama's price
  margin from the Intrade markets. Similarly, in a separate analysis,
  regress Obama's actual margin of victory on the Obama's predicted
  margin from the latest polls within each state. Interpret the
  results of these regressions.

## Answer 5

```{r}
## Regression from Intrade
int.fit08 <- lm(obama.actmarg ~ obama.intmarg, data = latest08)
coef(int.fit08)
## Regression from polls
## Subset latest poll predictions
poll.pred <- rep(NA, 51)
names(poll.pred) <- as.character(pres08$state)
for(i in 1:51){
    state.data <- subset(pollresults08, subset = (state == pres08$state[i]))
    ## Subset latest polls 
    ## Note alternative code using brackets for subsetting a dataframe
    latest <- state.data[state.data$DaysToElection == 
                             min(state.data$DaysToElection), ]
    ## Store predicted margins
    poll.pred[i] <- mean(latest$obama.pollmarg)
}
## Create dependent variable with the actual margin
pres08$obama.actmarg <- pres08$Obama - pres08$McCain
## Run regression
polls.fit08 <- lm(pres08$obama.actmarg ~ poll.pred)
coef(polls.fit08)
## Rsquare
summary(int.fit08)$r.squared
summary(polls.fit08)$r.squared
```

Both the Intrade predicted margin of victory for Obama and the latest
polls' predicted margin are positively associated with the actual
margin of victory for Obama across states.  That is, the actual margin
of victory for Obama is estimated to increase as Obama's predicted
margin increases.  In addition, based on the comparison of $R^2$
values, the polling model predicted the outcome better than Intrade.


## Question 6


Do the 2008 predictions of polls and Intrade accurately predict
  each state's 2012 elections results?  Using the fitted regressions
  from the previous question, forecast Obama's actual margin of
  victory for the 2012 election in two ways.  First, use the 2012
  Intrade price margins from the day before the election as the
  predictor in each state.  Recall that the 2012 Intrade data do not
  contain market prices for all states. Ignore states without data.
  Second, use the 2012 poll predicted margins from the latest polls in
  each state as the predictor, found in `polls12.csv`.

## Answer 6

We first compute the predictions based on the Intrade data.  We plot
the predicted state-wide election results against the actual
results. We also examine which states are misclassified.

```{r}
par(cex = 1.25)
## Compute 2012 forecast from Intrade
pred.int <- data.frame(obama.intmarg = latest12$obama.intmarg)
forecast.int <- predict(int.fit08, newdata = pred.int) 
names(forecast.int) <- latest12$state
## Plot the actual versus forecasted predictions
plot(forecast.int, latest12$obama.actmarg, type = "n", xlim = c(-50, 50),
     ylim = c(-50, 50), xlab = "Forecasted Obama Margin",
     ylab = "Actual Obama Margin by State",
     main = "Actual vs. Forecasted 2012 Results: Intrade")
text(forecast.int, latest12$obama.actmarg, latest12$state) 
abline(h = 0)
abline(v = 0)
## Which states misclassified?
latest12$state[sign(forecast.int) != sign(latest12$obama.actmarg)]
```

Next, we repeat the same analysis using the polling data. 

```{r}
par(cex = 1.25)
## Load 2012 Polls Data
polls12 <- read.csv("data/polls12.csv", header = T)
## Merge with 2012 results to get EV data
pollresults12 <- merge(polls12, pres12, by = "state")
## Calculate margins variable 
pollresults12$obama.pollmarg <-pollresults12$Obama.x - 
    pollresults12$Romney.x
## Create DaysToElection variable
pollresults12$DaysToElection <- as.Date("2012-11-06") - 
    as.Date(pollresults12$middate)
## Subset latest poll predictions
st.names12 <- unique(pollresults12$state) # To subset by state
length(st.names12) # number of states in the data
## loop starts here
poll.pred12 <- rep(NA, 47)
names(poll.pred12) <- as.character(st.names12)
for(i in 1:47){
  state.data <- subset(pollresults12, subset = (state == st.names12[i]))  
  ## Subset latest polls 
  latest <- state.data[state.data$DaysToElection == 
                           min(state.data$DaysToElection), ]
  ## Store predicted margins
  poll.pred12[i] <- mean(latest$obama.pollmarg)
}
## Compute 2012 forecast from polls
pred.polls <- data.frame(poll.pred = poll.pred12)
forecast.polls <- predict(polls.fit08, 
                          newdata = pred.polls) 
names(forecast.polls) <- names(poll.pred12)
## Plot the actual versus forecasted predictions
plot(forecast.polls, poll.pred12, type = "n", xlim = c(-50, 50),
     ylim = c(-50, 50), xlab = "Forecasted Obama Margin",
     ylab = "Actual Obama Margin by State",
     main = "Actual vs. Forecasted 2012 Results: Polls")
text(forecast.polls, poll.pred12, names(poll.pred12))
abline(h = 0)
abline(v = 0)
## Which states misclassified?
names(poll.pred12)[!is.na(forecast.polls) & sign(forecast.polls) != sign(poll.pred12)]
```

Both the Intrade and poll forecasts for 2012 predict the actual 2012
outcomes very accurately, although the Intrade forecasts for less
competitive states tend to exaggerate the actual margins in 2012. Both the
polling and Intrade forecasts for 2012 election accurately classify all states 
but Florida.
