---
title: "Success of Leader Assassination as a Natural Experiment (Graphs)"
output: html_document
---

One longstanding debate in the study of international relations
concerns the question of whether individual political leaders can make
a difference.  Some emphasize that leaders with different ideologies
and personalities can significantly affect the course of a nation.
Others argue that political leaders are severely constrained by
historical and institutional forces.  Did individuals like Hitler,
Mao, Roosevelt, and Churchill make a big difference?  The difficulty
of empirically testing these arguments stems from the fact that the
change of leadership is not random and there are many confounding
factors to be adjusted for.

In this exercise, we consider a *natural experiment* in which the
success or failure of assassination attempts is assumed to be
essentially random.

This exercise is based on:
Jones, Benjamin F, and Benjamin A Olken. 2009. “[Hit or Miss?
 The Effect of Assassinations on Institutions and
 War.](http://dx.doi.org/10.1257/mac.1.2.55)”
 *American Economic Journal: Macroeconomics* 1(2): 55–87.

Each observation of the CSV data set
`leaders.csv` contains information about an assassination
attempt.  The variables are:

-------------------------------------------------------------------------------
 Name                 Description
 -------------------- ----------------------------------------------------------
 `country`            The name of the country
 
 `year`               Year of assassination
 
 `leadername`         Name of leader who was targeted
 
 `age`                Age of the targeted leader
 
 `politybefore`       Average polity score during the 3 year period prior 
                      to the attempt
 
 `polityafter`        Average polity score during the 3 year period after 
                      the attempt
 
 `civilwarbefore`     1 if country is in civil war during the 3 year period prior 
                      to the attempt, or 0
 
 `civilwarafter`      1 if country is in civil war during the 3 year period after 
                      the attempt, or 0
 
 `interwarbefore`     1 if country is in international war during the 3 year period 
                      prior to the attempt, or 0
 
 `interwarafter`      1 if country is in international war during the 3 year period 
                      after the attempt, or 0
 
 `result`             Result of the assassination attempt, one of 10 categories described below
-------------------------------------------------------------------------------

The `polity` variable represents the so-called *polity score*
from the Polity Project.  The Polity Project systematically documents
and quantifies the regime types of all countries in the world from
1800.  The polity score is a 21-point scale ranging from -10
(hereditary monarchy) to 10 (consolidated democracy).   The
`result` variable is a 10 category factor variable describing
the result of each assassination attempt.

## Question 1

How many assassination attempts are recorded in the data?  How
  many countries experience at least one leader assassination attempt?
  (The `unique` function, which returns a set of unique values
  from the input vector, may be useful here).  What is the average
  number of such attempts (per year) among these countries?

## Answer 1
```{r}
leaders <- read.csv("data/leaders.csv")
## number of assassinations is number of rows
nrow(leaders)
## number of countries in the data
length(unique(leaders$country))
## average number of assassination attempts among these countries for
## each year, which then needs to be averaged across years
mean(tapply(leaders$country, leaders$year, length))
mean(table(leaders$year)) # alternative
```

The number of assassination attempts in the data is equal to the
number of observations, `r nrow(leaders)`. There are
`r length(unique(leaders$country))` countries in the dataset that
have experienced at least one assassination attempt. The average
number of attempts among all of these countries is equal to the
average number of observations per year in the data,
`r round(mean(table(leaders$year)), 2)`.

## Question 2

Create a new binary variable named `success` that is equal
  to 1 if a leader dies from the attack and to 0 if the leader
  survives.  Store this new variable as part of the original data
  frame.  What is the overall success rate of leader assassination?
  Does the result speak to the validity of the assumption that the
  success of assassination attempts is randomly determined?

## Answer 2

```{r}
## create variable, 1 if died "from" attack, 0 otherwise
lev <- levels(leaders$result)
lev # shows response categories
## exclude "dies for other reason" from successes
## because did not die from the attack
leaders$success <-
    ifelse(leaders$result == lev[1] | leaders$result == lev[2] |
               leaders$result == lev[3] | leaders$result == lev[4], 1, 0 )
## rate of success
mean(leaders$success)
```

Assassination attempts are successful
`r round(mean(leaders$success), 3)*100` percent of the time. It is
unclear whether this can validate the assumption that attempts are
randomly determined. It at least validates that not all attempts lead
to the death of the leader.  However, we do not know if other
systematic factors, such as skill of the perpetrator, health of the
leader, or security conditions, might be at work in determining
success and failure of the attempts.

## Question 3

Investigate whether the distribution of polity scores over 3 years prior to an assassination attempt differs between successful and failed attempts. To do this, plot the histograms for both distributions.  Assign informative labels to the axes and title your graphs. Add a vertical line to each graph to represent the mean values of the two distributions.  In addition, create a Quantile-Quantile plot and conduct the same comparison.  Briefly interpret the results in light of the validity of the aforementioned assumption.

## Answer 3

```{r}
par(cex = 1.25)
## histogram for successful attempts
hist(leaders$politybefore[leaders$success == 1], freq = FALSE,
     xlab = "Polity score", main = "Successful attempts",
     ylim = c(0, 0.12))
abline(v = mean(leaders$politybefore[leaders$success == 1]), lty = 3)
```

```{r}
## histogram for unsuccessful attempts
hist(leaders$politybefore[leaders$success == 0], freq = FALSE,
     xlab = "Polity score", main = "Unsuccessful attempts",
     ylim = c(0, 0.12))
abline(v = mean(leaders$politybefore[leaders$success == 0]), lty = 3)
```

Prior to the attempt, countries with unsuccessful assassinations were
on average as democratic as countries with successful attempts.  In
addition, both sets of countries have a greater number of less
democratic countries.  There are some differences in the two
distributions but it is difficult to know whether these are systematic
differences.  To conduct a more direct comparison, we create a
Quantile-Quantile plot.

```{r}
par(cex = 1.25)
qqplot(leaders$politybefore[leaders$success == 1],
       leaders$politybefore[leaders$success == 0],
       xlab = "Polity score prior to successful attempts",
       ylab = "Polity score prior to unsuccessful attempts")
abline(0, 1)
```
The Q-Q plot shows that the two distributions are largely identical,
suggesting that the success of assassination attempts is close to
random.

## Question 4

Investigate whether the success of an assassination attempt
varies by the country's experience of civil and international war.
  Create a new binary variable in the data frame called
  `warbefore`.  Code the variable such that it is equal to 1 if
  a country is in either civil or international war during the 3 years
  prior to an assassination attempt. Compare the average assassination rates
  for the two groups, and provide a brief interpretation
  of the result.

## Answer 4

```{r}
## create warbefore variable
table(leaders$interwarbefore)
table(leaders$civilwarbefore)
leaders$warbefore <- ifelse(leaders$interwarbefore == 1 |
                                leaders$civilwarbefore == 1, 1, 0)

## proportion war before successful/unsuccessful
mean(leaders$warbefore[leaders$success == 1])
mean(leaders$warbefore[leaders$success == 0])
## proportion of successful attempts
mean(leaders$success[leaders$warbefore == 1])
mean(leaders$success[leaders$warbefore == 0])
```

Countries with successful attempts were no more or less likely to be
engaged in war prior to the assassination attempt than those with
unsuccessful attempts.  This finding is consistent with the assumption
that the success of assassination attempts is random.

## Question 5

Does successful leader assassination cause democratization?
  Does successful leader assassination lead countries to war?  Answer
  these questions by analyzing the data. Be sure to state your
  assumptions and provide a brief interpretation of the results.

## Answer 5

```{r}
## compare polity scores before to scores after
diff.pol.succ <- mean(leaders$polityafter[leaders$success == 1]) -
    mean(leaders$politybefore[leaders$success == 1]) # among successful
diff.pol.unsucc <-mean(leaders$polityafter[leaders$success == 0]) -
    mean(leaders$politybefore[leaders$success == 0]) # unsuccessful
## difference in differences
diff.pol.succ - diff.pol.unsucc
```

To answer these questions, a difference-in-differences analysis is
conducted. We take the difference in mean polity scores before and
after the assassination attempts for countries with unsuccessful
attempts and subtract this difference from the difference score for
those with successful attempts. The assumption here is that the change
in polity scores for countries with successful attempts would be the
same as the change for those with unsuccessful attempts had their
assassinations not been successful.  The result is small and negative,
suggesting that successful attempts do not lead to democratization.

Note that if we had merely subtracted polity scores after the attacks
between successful and unsuccessful attempts, we would have seen a
positive difference and reached the opposite conclusion.

```{r}
## compare scores after
mean(leaders$polityafter[leaders$success == 1]) -
    mean(leaders$polityafter[leaders$success == 0])
```

Thus, taking into account differences in polity scores that existed
prior to the attempts was necessary for reaching an accurate
conclusion.  In contrast, because countries with and without
successful assassination attempts were no more or less likely to be
engaged in war prior to the attack, it is empirically less
consequential as to whether we account for pre-attempt factors to
answer the question. Using the same difference-in-difference approach,
we find very little difference in the proportion of countries engaged
in war. Leader assassination does not seem to cause countries to
engage in war.

```{r}
## create variable for warafter
leaders$warafter <- ifelse(leaders$interwarafter == 1 |
                               leaders$civilwarafter == 1, 1, 0)
## compare war before to war after among successful and unsuccessful
diff.war.succ <- mean(leaders$warafter[leaders$success == 1]) -
    mean(leaders$warbefore[leaders$success == 1])
diff.war.unsucc <-mean(leaders$warafter[leaders$success == 0]) -
    mean(leaders$warbefore[leaders$success == 0])
diff.war.succ - diff.war.unsucc # difference in differences
```


## Question 6

Consider the effect of leader assassination on democratization
  more closely.  Using a Quantile-Quantile plot, compare the changes
  in the distribution of polity scores among countries with successful
  attempts against changes in countries where assassinations failed.
  Be sure to use informative labels.  Provide a brief interpretation
  of the plot by connecting your analysis to the results obtained in
  the previous question.

## Answer 6

```{r}
## compare polity scores before to scores after, using full distributions
diff.pol.succ.dist <- leaders$polityafter[leaders$success == 1] -
    leaders$politybefore[leaders$success == 1] # among successful
diff.pol.unsucc.dist <- leaders$polityafter[leaders$success == 0] -
    leaders$politybefore[leaders$success == 0] # unsuccessful
```

Plot the two distributions against one another using a QQ plot:
```{r}
par(cex = 1.25)
## plot the distributions using a QQ plot
qqplot(diff.pol.succ.dist, diff.pol.unsucc.dist,
       xlim = c(-15, 15), ylim = c(-15, 15),
       main = "Prospects of Democtratization following Assassinations
       (measuring difference in polity scores)",
       xlab = "Successful assassinations", ylab = "Unsuccessful assassinations")
## add a 45-degree line
abline(0, 1, lty = 3)
```

Notice that the axes show the magnitude of the change in a country's
polity score before and after an assassination. Notice that the two
distributions diverge mostly for lower differences in polity scores,
roughly in the range (-2,-15).  However, most of the quantiles of the
two distributions align closely, as indicated by their proximity to
the 45-degree line. These two observations confirm the results
obtained in the previous question: we can discern a negative effect of
successful assasination on the likelihood of democratization, but the
effect appears to be subtle, not affecting the center of the
distribution.



