---
title: "Effect of Demographic Change on Exclusionary Attitudes"
output:
  pdf_document: default
  html_document: default
---

A researcher conducted a randomized field experiment assessing the
extent to which individuals living in suburban communities around
Boston, Massachusetts, and their views were affected by exposure to
demographic change. 

This exercise is based on: Enos, R. D. 2014. “[Causal Effect of 
Intergroup Contact on Exclusionary 
Attitudes.](http://dx.doi.org/10.1073/pnas.1317670111)” 
*Proceedings of the National Academy of Sciences* 111(10): 3699–3704.

Subjects in the experiment were individuals riding on the commuter
rail line and overwhelmingly white. Every morning, multiple trains
pass through various stations in suburban communities that were used
for this study. For pairs of trains leaving the same station at
roughly the same time, one was randomly assigned to receive the
treatment and one was designated as a control. By doing so all the
benefits of randomization apply for this dataset.

The treatment in this experiment was the presence of two native Spanish-speaking 
'confederates' (a term used in experiments to indicate that these
individuals worked for the researcher, unbeknownst to the subjects)
 on the platform each morning prior to the train's arrival. The
presence of these confederates, who would appear as Hispanic foreigners 
to the subjects,  was intended to
simulate the kind of demographic change anticipated for the United
States in coming years. For those individuals in the control group, no
such confederates were present on the platform. The treatment was
administered for 10 days.  Participants were asked questions related
to immigration policy both before the experiment started and after the
experiment had ended.  The names and
descriptions of variables in the data set `boston.csv` are:

 -------------------- ----------------------------------------------------------
 Name                 Description
 -------------------- ----------------------------------------------------------
 `age`                Age of individual at time of experiment 

 `male`               Sex of individual, male (1) or female (0) 

 `income`             Income group in dollars (not exact income) 

 `white`              Indicator variable for whether individual 
                      identifies as white (1) or not (0) 
                      
 `college`            Indicator variable for whether individual 
                      attended college (1) or not (0) 

 `usborn`             Indicator variable for whether individual is 
                      born in the US (1) or not (0) 

 `treatment`          Indicator variable for whether an individual 
                      was treated (1) or not (0) 

 `ideology`           Self-placement on ideology spectrum from Very Liberal (1) 
                      through Moderate (3) to Very Conservative (5) 

 `numberim.pre`       Policy opinion on question about increasing the number 
                      immigrants allowed in the country from Increased (1) to Decreased (5)

 `numberim.post`      Same question as above, asked later

 `remain.pre`         Policy opinion on question about allowing the children of 
                      undocumented immigrants to remain in the country from 
                      Allow (1) to Not Allow (5) 

 `remain.post`        Same question as above, asked later

 `english.pre`        Policy opinion on question about passing a law establishing 
                      English as the official language from Not Favor (1) to Favor (5) 

 `english.post`       Same question as above, asked later
 
 ------------------------------------------------------------------------------
 
## Question 1

The benefit of randomly assigning individuals to the treatment
  or control groups is that the two groups should be similar, on
  average, in terms of their covariates. This is referred to as
  'covariate balance.' Show that the treatment and control groups
  are balanced with respect to the income variable (`income`) by
  comparing its distribution between those in the treatment group and
  those in the control group.  Also, compare the proportion of males
  (`male`) in the treatment and control groups. Interpret these
  two numbers.

## Answer 1

```{r}
## load data
boston <- read.csv("data/boston.csv", head = TRUE)
## create two subsets
boston.treat <- subset(boston, treatment == 1)
boston.ctrl <- subset(boston, treatment == 0)
summary(boston.treat$income)
summary(boston.ctrl$income)  
median(boston.treat$income)
median(boston.ctrl$income)
median(boston.ctrl$income) - median(boston.treat$income)
```
    
The median income for each group is the same. So, the distribution
of household incomes across the control and treatment group are
balanced.

```{r}
mean(boston.treat$male)
mean(boston.ctrl$male)
```

Of all the individuals assigned to the treatment group,
`r round(mean(boston.treat$male),3)*100`% are males. Of all the
individuals assigned to the control group,
`r round(mean(boston.ctrl$male),3)*100`%
are male. They differ by about
`r round((mean(boston.ctrl$male) -
  mean(boston.treat$male))*100,0)`
percentage points. The treatment and control groups are relatively
balanced on income and gender.  

## Question 2
      
Individuals in the experiment were asked a series of questions
  both at the beginning and the end of the experiment. One such
  question was "Do you think the number of immigrants from Mexico who
  are permitted to come to the United States to live should be
  increased, left the same, or decreased?" The response to this
  question prior to the experiment is in the variable
  `numberim.pre`. The response to this question after the
  experiment is in the variable `numberim.post`. In both cases
  the variable is coded on a 1 -- 5 scale. Responses with values of 1
  are inclusionary ('pro-immigration') and responses with values of
  5 are exclusionary ('anti-immigration').  Compute the average
  treatment effect on the change in attitudes about immigration. That
  is, how does the mean change in attitudes about immigration policy
  for those in the control group compare to those in the treatment
  group. Interpret the result.

## Answer 2

```{r}
## compute change in attitude
boston$change <- boston$numberim.post - boston$numberim.pre
treat.change <- mean(boston$change[boston$treatment == 1],
                     na.rm = TRUE)
ctrl.change <- mean(boston$change[boston$treatment == 0],
                    na.rm = TRUE)
treat.change
ctrl.change
## average treatment effect
treat.change - ctrl.change
```
    
The average change within the control group is a decrease in
`r round(ctrl.change*-1, 2)` points on the 5 point scale (where a decrease
indicates more inclusionary attitudes). The average change within the
treatment group is an increase in `r round(treat.change, 2)` points on the 5
point scale (where an increase indicates more exclusionary
attitudes). The average treatment effect is the difference of the
two. On average, the responses of individuals in the treatment group
were more exclusionary than the control group by
`r round(treat.change - ctrl.change,2)` points. Exposure to
simulated demographic changes caused this increase in exclusionary
attitudes.

## Question 3

Does having attended college influence the effect of being
  exposed to 'outsiders' on exclusionary attitudes? Another way to
  ask the same question is this: is there evidence of a differential
  impact of treatment, conditional on attending college versus not
  attending college?  Calculate the necessary quantities to answer
  this question and interpret the results. Consider the average
  treatment effect for those who attended college and then those who
  did not.

## Answer 3

```{r}
t.col.change <- mean(boston$change[boston$treatment == 1 &
                                       boston$college == 1],
                     na.rm = TRUE)
c.col.change <- mean(boston$change[boston$treatment == 0 &
                                       boston$college == 1],
                     na.rm = TRUE)
t.nocol.change <- mean(boston$change[boston$treatment == 1 &
                                         boston$college == 0],
                       na.rm = TRUE)
c.nocol.change <- mean(boston$change[boston$treatment == 0 &
                                         boston$college == 0],
                       na.rm = TRUE)
ate.col <- t.col.change - c.col.change
ate.nocol <- t.nocol.change - c.nocol.change
ate.col
ate.nocol
ate.col - ate.nocol
```
    
The average treatment effect (using the 'change' variable) among
those with a college education is an increase in exclusionary
attitudes of about `r round(ate.col,2)` points. Among those
without a college education, there is an increase in exclusionary
attitudes of about `r round(ate.nocol,2)`. Both of these effects
are on a 5 point scale.

The difference between them is less than `r round(ate.col -
  ate.nocol,2)` points on this scale and is negligible. Because the
average treatment effects do not vary across the two subgroups, there
is no evidence of a differential impact.

## Question 4

Repeat the same analysis as in the previous question but this
  time with respect to age and ideology.  For age, divide the data
  based on its quartile and compute the average treatment effect
  within each of the resulting four groups.  For ideology, compute the
  average treatment effect within each value.  What patterns do you
  observe?  Give a brief substantive interpretation of the results.
    
## Answer 4

```{r}
## create age quartiles
qrt <- quantile(boston$age, probs = seq(from = 0, to = 1, by = 0.25))
boston$age.qrt <- 
    ifelse(boston$age <= qrt[2], 1,
           ifelse(boston$age > qrt[2] & boston$age <= qrt[3], 2,
                  ifelse(boston$age > qrt[3] & boston$age <= qrt[4], 3, 
                         ifelse(boston$age > qrt[4], 4, NA))))
## treatmemnt group change by age
t.age.change <- tapply(boston$change[boston$treatment == 1], 
                       boston$age.qrt[boston$treatment == 1], mean, 
                       na.rm = TRUE)
## control group change by age
c.age.change <- tapply(boston$change[boston$treatment == 0], 
                       boston$age.qrt[boston$treatment == 0], mean, 
                       na.rm = TRUE)
ate.age <- t.age.change - c.age.change
ate.age

## treatment group change by ideology
t.idea.change <- tapply(boston$change[boston$treatment == 1],
                        boston$ideology[boston$treatment == 1],
                        mean, na.rm = TRUE)
## control group change by ideology
c.idea.change <- tapply(boston$change[boston$treatment == 0],
                        boston$ideology[boston$treatment == 0],
                        mean, na.rm = TRUE)
ate.idea <- t.idea.change - c.idea.change
ate.idea
```
    
Within each age category and ideology category, we see the largest
changes between treatment and control groups at the extremes of the
distribution. We also see a fair amount of difference in treatment
effects, suggesting there may be a differential impact by age and by
ideology.

    

