---
title: "Predicting the 2016 US Presidential Election"
output:
  pdf_document: default
  html_document: default
---

The U.S. president is elected by the electoral college -- 538 electors
corresponding to 435 members of congress, 100 senators, and 3
additional electors allocated to Washington D.C.. The number of
electoral votes allocated to each state is equal to the size of its
congressional delegation. And most states cast all their electoral
votes for the candidate receiving a plurality of the state's votes in
the general election (the *winner-takes-all* rule). Nebraska and Maine
are the only two exceptions. These states allocate two electoral votes
to the candidate receiving a plurality of the state's votes, and each
of their remaining electoral votes go to the candidate receiving a
plurality of votes within each of the states' congressional
districts. But these are small and relatively homogeneous
states. Maine has never actually split its electoral votes and
Nebraska did it only once, casting a vote for Obama in 2008.

A candidate must receive a simple majority of electoral college votes
(270 votes) to be elected. But, as we have seen in 2000, it is
possible for a candidate to win the election without receiving a
plurality of the popular vote. In today's precept we will analyze
state-level polls downloaded from the Huffington Post's Pollster
(http://elections.huffingtonpost.com/pollster/polls) and 3 additional
polls for Washington D.C. available at
(http://www.electoral-vote.com/evp2016/Pres/pres_polls.txt) to predict
the outcomes of the 2016 presidential election. We will predict the
distribution of electoral college votes according to the
*winner-takes-all* rule and using only the 3 most recent polls in each
state and examine how this distribution changed over time, starting at
90 days before the election.

The dataset we will be using this week (`polls2016.csv`) has 
`r nrow(read.csv("data/polls2016.csv"))` observations, each representing
a different poll, and includes the following 
`r ncol(read.csv("data/polls2016.csv"))` variables:

-----------------------------------------------------------------------------
Name                Description
---------------     ---------------------------------------------------------
`id`                Poll ID

`state`             U.S. state where poll was fielded

`Clinton`           The poll's estimated level of support for Hillary Clinton   
                    (in percentage points) 

`Trump`             The poll's estimated level of support for Donald Trump (in 
                    percentage points)

`days_to_election`  Number of days before November 4, 2016. 

`electoral_votes`   Number of electoral votes allocated to the state where the poll was 
                    fielded (a state-level variable)
                    
`population`        The poll's target population, which may be `Adults`, `Registered Voters`, 
                    or `Likely Voters` 
--------------------------------------------------------------------------------

## Question 1

We will begin by restricting our poll data to the 3 most recent polls
in each state and computing the average support for each candidate by
state. Create a scatterplot showing support for Clinton vs. support
for Trump. Use state abbreviations to plot the results. Briefly
interpret the results.

**Hint:** To do this see the code in Section 4.1.3 of QSS. The only
difference is that you will have to sort the polls by the
`days_to_election` variable within each state.  Use the `sort()`
function to sort the polls from the latest to the oldest.  When the
`index.return` argument is set to `TRUE`, this function will return
the ordering index vector, which can be used to extract the 3 most
recent polls for each state.

## Answer 1

```{r}
## load data:
polls <- read.csv("data/polls2016.csv")
## state names
state.names <- unique(polls$state)
## initialize prediction vectors
Clinton.support <- Trump.support <- rep(NA, 51)
names(Clinton.support) <- names(Trump.support) <- state.names
for(i in 1:51){
    state.data <- subset(polls, state == state.names[i]) # subset each state
    idx <- sort(state.data$days_to_election, index.return = TRUE) # sort by date
    state.recent <- state.data[idx$ix[1:3], ]  # use sorted index to get the 3 latest polls
    Clinton.support[i] <- mean(state.recent$Clinton)
    Trump.support[i] <- mean(state.recent$Trump)
}
## scatterplot
plot(Trump.support, Clinton.support, type='n', xlim = c(15,80), ylim = c(15,80),
     xlab = "Trump's Support", ylab = "Clinton's Support")
## add state abbreviations instead of points
text(Trump.support, Clinton.support, labels = state.names)
## 45-degree line
abline (a = 0, b = 1, lty = "dashed")
```

The scatterplot shows support for Clinton on the y-axis and Support
for Trump on the x-axis. States where Trump is expected to win appear
below the 45 degree line, which is indicated by a dashed line, and
states where Clinton is expected to win appear above this line. States
where both candidates have nearly the same level of support (see
Arizona) appear along the 45 degree line. These are the potentially
contested states. Nearly perpendicular to the 45 degree line we see
the expected negative relationship between support for Clinton and
Support for Trump. Utah is a notable outlier because Trump is only a
few points ahead of Clinton there and neither candidate is broadly
supported. Instead, an indepenent candidate, Evan McMullin, is doing
well.

## Question 2

Based on the average support you calculated for Clinton and Trump,
predict the winner of each state and allocate the corresponding
electoral college votes to the predicted winner. While two states,
Maine and Nebraska, do not apply the *winner-takes-all* rule to
allocate their electoral votes, for the sake of simplicity, we will
apply this rule uniformly across these states as well. If the support
for the two candidates in a given state is identical, split the
state's electoral votes. Who do you predict will win the election? How
many electoral college votes do you predict each candidate will
receive?

## Answer 2

We can do this with a loop, which is intuitive:

```{r}
## Electoral votes for each state
ev <- tapply(polls$electoral_votes, polls$state, mean)
Clinton <- Trump <- 0 # candidates' electoral votes
for (i in 1:length(ev)) {
  if (Trump.support[i] > Clinton.support[i]) { # Trump wins
    Trump <- Trump + ev[state.names[i]]
  } else if (Trump.support[i] < Clinton.support[i]) { # Clinton wins
    Clinton <- Clinton + ev[state.names[i]]
  } else { # a tie
    Clinton <- Cliinton + ev[state.names[i]] / 2
    Trump <- Trump + ev[state.names[i]] / 2
  }  
}
names(Clinton) <- "Cinton EV"
names(Trump) <-"Trump EV"
## total predicted eletoral votes
Clinton
Trump
```

Alternatively, we can do it without a loop, which is a little trickier:

```{r}
## sort electoral votes to the order states appear in the data 
ev.sorted <- ev[state.names]
## Clinton's electoral votes
sum(ev.sorted[Clinton.support > Trump.support])
## Trump's electoral votes
sum(ev.sorted[Clinton.support < Trump.support])
## check to see if there is any tie
sum(Clinton.support == Trump.support)
```

Using only the 3 most recent polls for each state, we predict that
Clinton will be elected president, receiving `r Clinton` electoral
votes compared to Trump's `r Trump`.

## Question 3

Let's examine how our predictions may have differed if we had used
only polls based on *likely voters*.  Since we have fewer polls that
are based on likely voters, for each state compute the average of the
most recent poll (based on the `days_to_election` variable) and those
conducted within 30 days from it.  In addition, assume that Clinton
will win Washington DC. How does the result change when compared to
Question 2?  Repeat the question but this time using the polls based
on *registered voters*.  Briefly interpret the results.

## Answer 3

```{r}
## Subset poll data
likely <- subset(polls, population == "Likely Voters")
registered <- subset(polls, population == "Registered Voters")
## electoral college votes
Clinton.likely <- Clinton.registered <- Trump.likely <- Trump.registered <- 0
for (i in 1:51){
    if (state.names[i] == "DC") {
      Clinton.likely <- Clinton.likely + ev["DC"]
      Clinton.registered <- Clinton.registered + ev["DC"]
    } else {
      ## subset data for state
      likely.state <- subset(likely, state == state.names[i])
      registered.state <- subset(registered, state == state.names[i])
      ## date of the latest poll within state
      latest.likely <- min(likely.state$days_to_election)
      latest.registered <- min(registered.state$days_to_election)
      ## further subset the polls
      polls.likely <- subset(likely.state, days_to_election <= (latest.likely + 30))
      polls.registered <- subset(registered.state, days_to_election <= (latest.registered + 30))
      ## compute the average support
      C.likely <- mean(polls.likely$Clinton)
      T.likely <- mean(polls.likely$Trump)
      C.registered <- mean(polls.registered$Clinton)
      T.registered <- mean(polls.registered$Trump)
      ## electoral college votes allocation
      if (C.likely > T.likely) {
        Clinton.likely <- Clinton.likely + ev[state.names[i]]
      } else if (C.likely < T.likely) {
        Trump.likely <- Trump.likely + ev[state.names[i]]
      } else {
        Clinton.likely <- Clinton.likely + ev[state.names[i]] / 2
        Trump.likely <- Trump.likely + ev[state.names[i]] / 2
      }
      if (C.registered > T.registered) {
        Clinton.registered <- Clinton.registered + ev[state.names[i]]
      } else if (C.registered < T.registered) {
        Trump.registered <- Trump.registered + ev[state.names[i]]
      } else {
        Clinton.registered <- Clinton.registered + ev[state.names[i]] / 2
        Trump.registered <- Trump.registered + ev[state.names[i]] / 2
      }
    } 
}
names(Clinton.likely) <- "Clinton, likely voters"
names(Trump.likely) <- "Trump, likely voters"
names(Clinton.registered) <- "Clinton, registered voters"
names(Trump.registered) <- "Trump, registered voters"
## likely voters
Clinton.likely
Trump.likely
## registered voters
Clinton.registered
Trump.registered
```

Both among likely voters and among registered voters, the number of
electoral votes predicted for Clinton is higher when we used the most
recent polls than when we averaged the three most recent polls in
Question 2. Comparing the latest polls of likely voters to the latest
polls of registered voters, we see that the polls of likely voters
were less favorable to Clinton than the polls of registered
voters. Using the latest polls of likely voters we predicted that
Clinton would receive `r Clinton.likely` electoral votes vs. `r
Trump.likely` for Trump. Restricting our analysis to polls of
registered voters, we predicted that Clinton would receive `r
Clinton.registered` electoral votes versus `r Trump.registered` for
Trump.

## Question 4

Finally, we examine how poll predictions have changed over the past
few weeks. Starting at 60 days before the election, and for each day,
repeat the same analysis as the one conducted for Question 1. That is,
for each day, we take the 3 latest polls (or fewer if only one or two
is available) for each state and compute the average support
separately for Clinton and Trump within each state. We then allocate
the electoral votes of that state based on its predicted winner. Use a
time series plot to present the predicted total number of electoral
votes for each candidate. Add the winning line, i.e., an absolute
majority of 270 votes, as a horizontal line. Briefly describe the
results.

**Hint:** You may want to create a nested loop, in which an outer loop
is for each day, starting 60 days before the election and stopping
on the day of the most recent poll, and an inner loop is used to
calculate support for each candidate by state using the 3 most
recent polls on any given day.

## Answer 4

```{r}
## number of days for prediction
days <- 60 - min(polls$days_to_election) + 1
## set-up prediction vectors
Clinton <- Trump <- rep(0, days)
for (i in 1:days){ # loop for days
    polls.day <- subset(polls, days_to_election >= (60 - i + 1))
    for (j in 1:51){ # loop for states
        state.data <- subset(polls.day, state == state.names[j])
        idx <- sort(state.data$days_to_election, index.return = TRUE) # sort by date
        state.recent <- state.data[idx$ix[1:3], ]  # use sorted index to get the 3 latest polls
        Clinton.state <- mean(state.recent$Clinton, na.rm = TRUE)
        Trump.state <- mean(state.recent$Trump, na.rm = TRUE)
        if (Clinton.state > Trump.state) {
          Clinton[i] <- Clinton[i] + ev[state.names[j]]
        } else if (Clinton.state < Trump.state) {
          Trump[i] <- Trump[i] + ev[state.names[j]]
        } else {
          Clinton[i] <- Clinton[i] + ev[state.names[j]] / 2
          Trump[i] <- Trump[i] + ev[state.names[j]] / 2
        }
    }
}

## plot time series from 60 days to date of last polls
plot(60:min(polls$days_to_election), Clinton, type = 'l', xlim = c(60, 0), ylim = c(100, 400), 
     col = 'blue', xlab = 'Days to the election', ylab = 'Electoral votes')
lines(60:min(polls$days_to_election), Trump, type = 'l', col = 'red')
## add a line for 270 electoral votes
abline(h = 270)
## label candidates
text(50, 160, 'Trump', col = 'red')
text(50, 360, 'Clinton', col = 'blue')
```

The graph shows how the predicted number of electoral votes for each
candidate changed over the last `r days` days, starting at 60 days
before the election. On each day the number of electoral votes for
each candidate was predicted by the 3 most recent available polls in
each state (or fewer polls when 3 polls were not yet
available). Results above the horizontal line at 270 electoral votes
show which candidate is predicted to win the election. The number of
electoral votes predicted for each candidate changed over the 42 days
we analyzed, but support for Clinton has not dropped below the 270
votes line.
