---
title: "Poverty and Economic Decision-Making"
output:
  pdf_document: default
  html_document: default
---

Do changes in one's financial circumstances affect one's decision-making process and cognitive capacity? In an experimental study, researchers randomly selected a group of US respondents to be surveyed before their payday and another group to be surveyed after their payday. Under this design, the respondents of the `Before Payday` group are more likely to be financially strained than those of the `After Payday` group.  The researchers were interested in investigating whether or not changes in people's financial circumstances affect their decision making and cognitive performance.  Other researchers have found that scarcity induce an additional mental load that impedes cognitive capacity.  This exercise is based on:

Carvalho, Leandro S., Meier, Stephen, and Wang, Stephanie W. (2016). "[Poverty and economic decision-making: Evidence from changes in financial resources at payday.](http://dx.doi.org/10.1257/aer.20140481)" *American Economic Review*, Vol. 106, No. 2, pp. 260-284. 

In this study, the researchers administered a number of decision-making and cognitive performance tasks to the `Before Payday` and `After Payday` groups. We focus on the *numerical stroop task*, which measures cognitive control.  In general, taking more time to complete this task indicates less cognitive control and reduced cognitive ability. They also measured the amount of cash the respondents have, the amount in their checking and saving accounts, and the amount of money spent.  The data set is in the CSV file `poverty.csv`. The names and descriptions of variables are given below:

-------------------------------------------------------------------------------
 Name                         Description
 ---------------------------- ---------------------------------------------------------
  `treatment`                 Treatment conditions: `Before Payday` and `After Payday`
  
  `cash`                      Amount of cash respondent has on hand
  
  `accts_amt`                 Amount in checking and saving accounts
  
  `stroop_time`               Log-transformed average response time for cognitive 
                              stroop test
  
  `income_less20k`            Binary variable: `1` if respondent earns less than 20k a 
                              year and `0` otherwise
-------------------------------------------------------------------------------
 
## Question 1

Load the `poverty.csv` data set. Look at a summary of the `poverty` data set to get a sense of what its variables looks like. Use histograms to examine the univariate distributions of the two financial resources measures: `cash` and `accts_amt`. What can we tell about these variables' distributions from looking at the histograms? Evaluate what the shape of these distributions could imply for the authors' experimental design.
 
Now, take the *natural logarithm* of these two variables and plot the histograms of these tranformed variables. How does the distribution look now? What are the advantages and disadvantages of transforming the data in this way? 

**NOTE:** Since the natural logarithm of 0 is undefined, researchers often add a small value (in this case, we will use \$1 so that $\log 1 = 0$) to the 0 values for the variables being transformed (in this case, `cash` and `accts_amt`) in order to successfully apply the `log()` function to all values. Be sure to do this recoding only for the purposes of taking the logarithmic transformation -- keep the original variables the same. 

## Answer 1

```{r}
poverty <- read.csv("data/poverty.csv")
hist(poverty$cash, freq = FALSE, xlab = "Cash", main = "Distribution of Cash")

hist(poverty$accts_amt, freq = FALSE, xlab = "Checking/Savings Balance", 
     main = "Distribution of Checking/Savings Balance")

```

The distributions of the `cash`, and `accts_amt` variables are all highly skewed and possess many extreme values (outliers). These extreme values could pull our mean estimates upward, causing us to overestimate the mean amount of cash, as well as amount in checking and saving accounts for our sample. If these outliers are not evenly distributed between `BeforeTreatment` and `AfterTreatment` groups, we could also mischaracterize the effect of change in financial resources on decision-making. This would hamper the authors' ability to make a causal claim about this effect by hurting the internal validity of the survey experiment.


```{r}
poverty$log_cash <- poverty$cash
poverty$log_cash[poverty$log_cash == 0] <- 1
poverty$log_cash <- log(poverty$log_cash)

poverty$log_accts_amt <- poverty$accts_amt
poverty$log_accts_amt[poverty$log_accts_amt == 0] <- 1
poverty$log_accts_amt <- log(poverty$log_accts_amt)
  
hist(poverty$log_cash, freq = FALSE, xlab = "log(Cash)",
     main = "Distribution of log(Cash) After Payday")

hist(poverty$log_accts_amt, freq = FALSE, 
     xlab = "log(Checking/Savings Balance)",
     main = "Distribution of log(Checking/Savings)\n Balance After Payday")

```

Taking the natural log of the three variables decreased the amount of variation, causing the histograms to look much less skewed. However, it is harder to interpret the values of the variables since everything is now on a log scale. 

## Question 2

Now, let's examine the primary outcome of interest for this study-- the effect of a change in financial situation (in this case, getting paid on payday) on economic decision-making and cognitive performance. Begin by calculating the treatment effect for the `stroop_time` variable (a log-transformed variable of the average response time for the stroop cognitive test), using first the mean and then the median. What does this tell you about differences in the outcome across the two experimental conditions? 

Secondly, let's look at the relationship between finanical circumstances and the cognitive test variable. Produce two scatter plots side by side (hint: use the par(mfrow)) before your plot commands to place graphs side-by-side), one for each of the two experimental conditions, showing the bivariate relationship between your *log-transformed* `cash` variable and the amount of time it took subjects to complete the stroop cognitive test administered in the survey (`stroop_time`). Place the `stroop_time` variable  on the y-axis. Be sure to title your graphs to differentiate between the `Before Payday` and `After Payday` conditions. Now do the same, for the *log-transformed* `accts_amt` variable. 

Briefly comment on your results in light of the hypothesis that changes in economic circumstances will influence cognitive performance. 

## Answer 2

```{r}

tapply(poverty$stroop_time, poverty$treatment, mean) - 
  mean(poverty$stroop_time[poverty$treatment == "Before Payday"])

tapply(poverty$stroop_time, poverty$treatment, median) -
  median(poverty$stroop_time[poverty$treatment == "Before Payday"])

```

Because mean and median cognitive scores are very similar for the two groups, the average and median treatment effects are close to zero. 

```{r}
par(mfrow = c(1,2))
plot(poverty$log_cash[poverty$treatment == "Before Payday"], 
     poverty$stroop_time[poverty$treatment == "Before Payday"],
     ylab = "log(Average Response Time)", xlab = "log(Cash)", main = "Before Payday")

plot(poverty$log_cash[poverty$treatment == "After Payday"], 
     poverty$stroop_time[poverty$treatment == "After Payday"],
     ylab = "log(Average Response Time)", xlab = "log(Cash)", main = "After Payday")

plot(poverty$log_accts_amt[poverty$treatment == "Before Payday"], 
     poverty$stroop_time[poverty$treatment == "Before Payday"],
     ylab = "log(Average Response Time)", xlab = "log(Amount in Accounts)", main ="Before Payday")

plot(poverty$log_accts_amt[poverty$treatment == "After Payday"], 
     poverty$stroop_time[poverty$treatment == "After Payday"],
     ylab = "log(Average Response Time)", xlab = "log(Amount in Accounts)", 
     main = "After Payday")
```

By looking at the scatter plots we can see that the relationship between financial circumstances (both cash on hand and amount in accounts), and cognitive performance (as indicated by the amount of time it took respondents to do the stroop test) is essentially flat for both before and after payday groups. This suggests that changes in financial resources do *not* affect cognitive performance. 

## Question 3

Now, let's take a closer look at whether or not the `Before Payday` versus `After Payday` treatment  created measurable differences in financial circumstances. What is the effect of payday on participants' financial resources? To  help with interpretability, use the original variables `cash` and `accts_amt` to calculate this effect. Calculate both the mean and median effect. Does the measure of central tendency you use affect your perception of the effect?

## Answer 3

```{r}
cashmean.diff <- mean(poverty$cash[poverty$treatment == "After Payday"], na.rm = TRUE) - 
  mean(poverty$cash[poverty$treatment == "Before Payday"], na.rm = TRUE)

cashmed.diff <- median(poverty$cash[poverty$treatment == "After Payday"], na.rm = TRUE) - 
  median(poverty$cash[poverty$treatment == "Before Payday"], na.rm = TRUE)

cash.diff <- c(cashmean.diff, cashmed.diff)
names(cash.diff) <- c("Mean Diff.in Cash", "Median Diff.in Cash")
cash.diff
```

After payday, respondents report on average `r round(cashmean.diff, 2)` dollars more in cash holding (cash they had on hand) than Before payday respondents. However, the median difference in amount of cash reported is `r round(cashmed.diff, 2)` dollars. Since the mean is more sensitive to outliers, this relatively small difference between the two measures of central tendency is not surprising.

```{r}
balancemean.diff <- mean(poverty$accts_amt[poverty$treatment == "After Payday"], na.rm = TRUE) - 
  mean(poverty$accts_amt[poverty$treatment == "Before Payday"], na.rm = TRUE)

balancemed.diff <- median(poverty$accts_amt[poverty$treatment == "After Payday"], na.rm = TRUE) - 
  median(poverty$accts_amt[poverty$treatment == "Before Payday"], na.rm = TRUE)

balance.diff <- c(balancemean.diff, balancemed.diff)
names(balance.diff) <- c("Mean Diff.in Balance", "Median Diff.in Balance")
balance.diff
```

After payday, respondents report on average `r round(-balancemean.diff, 2)` dollars *less* in their checking and savings than before payday. The median difference, however, indicates that respondents have `r round(balancemed.diff, 2)` *more* dollars in their checking and savings account. This large diffence between the two values suggests the presence of influential outliers unevenly distributed between before and after payday groups.

## Question 4

Compare the distributions of the `Before Payday` and `After Payday` groups for the *log-transformed* `cash` and `accts_amt` variables. Use quantile-quantile plots to do this comparison, and add a 45-degree line in a color of your choice (not black). Briefly interpret your results and their implications for the authors' argument that their study generated variation in financial resources before and after payday. When appropriate, state which ranges of the outcome variables you would focus on when comparing decision-making and cognitive capacity across these two treatment conditions. 

## Answer 4

```{r}
par(cex = 1.25)
qqplot(poverty$log_cash[poverty$treatment == "Before Payday"], 
       poverty$log_cash[poverty$treatment == "After Payday"],
       xlab = "Before Payday", ylab = "After Payday",
       main = "QQ-plot Logged Cash Variable:\n Before vs. After Payday")
abline(0,1, col="red")
```

Despite the presence of some outliers, the qq-plot is relatively close to a 45 degree line. This suggests that there are no major differences in the distribution of cash held by respondents `Before Payday` and `After Payday`. This could be a problem for the authors' argument because cash is one of the measures of financial resources the authors use. If there is no variation in cash after payday anywhere in the distribution, the treatment did not occur.

```{r}
par(cex = 1.25)
qqplot(poverty$log_accts_amt[poverty$treatment == "Before Payday"], 
       poverty$log_accts_amt[poverty$treatment == "After Payday"],
       xlab = "Before Payday", ylab = "After Payday",
       main = "QQ-plot Logged Account Balance Variable:\n Before vs. After Payday")
abline(0, 1, col = "red")
```

While after payday respondents in the lower 50 - 75 quantiles seem to have more money in checking and savings accounts than those before payday, the upper 25 - 50 quantiles of the qq-plot look relatively close to a 45 degree line. If these differences are systematically associated with a pre-treatment variable, this would suggest that there is a heterogeneous effect of payday on checking and savings account balances. This could be a problem for the authors' argument because checking and savings account balance is one of the measures of financial resources they use. If there is no difference in checking and savings account balances for `Before Payday` and `After Payday`, this means that treatment did not take. Given this heterogeneous treatment effect, we can only test the authors' hypotheses about the relationship between changes in financial resources and decision-making/cognitive performance for some of the respondents --- those most likely to be at the lower end of the `accts_amt` variable's distribution. Unfortunately, if those respondents are systematically different from respondents most likely to be at the upper end of the accts_amt variable's distribution, this could put the study's internal validity in question.

## Question 5 

In class, we covered the difference-in-difference design for comparing average treatment effects across treatment and control groups. This design can also be used to compare average treatment effects across different ranges of a *pre-treatment variable*- a variable that asks about people's circumstances before the treatment and thus could not be affected by the treatment. This is known as *heterogeneous treatment effects* -- the idea that the treatment may have differential effects for different subpopulations. Let's look at the pre-treatment variable `income_less20k`. Calculate the treatment effect of Payday on amount in checking and savings accounts separately for respondents earning more than 20,000 dollars a year and those earning less than 20,000 dollars. Use the original `accts_amt` variable for this calculation. Then take the difference between the effects you calculate. What does this comparison tell you about how payday affects the amount that people have in their accounts? Are you convinced by the authors' main finding from Question 2 in light of your investigation of their success in manipulating cash and account balances before and after payday?

## Answer 5 

```{r}
great20k_diff <- mean(poverty$accts_amt[poverty$income_less20k == 0 & 
                                          poverty$treatment == "After Payday"], na.rm = TRUE) -
  mean(poverty$accts_amt[poverty$income_less20k == 0 & 
                           poverty$treatment == "Before Payday"], na.rm = TRUE)

less20k_diff <- mean(poverty$accts_amt[poverty$income_less20k == 1 & 
                                         poverty$treatment == "After Payday"], na.rm = TRUE) -
  mean(poverty$accts_amt[poverty$income_less20k == 1 & 
                           poverty$treatment == "Before Payday"], na.rm = TRUE)
income_diffs <- c(great20k_diff, less20k_diff)
names(income_diffs) <- c("income > 20,000", "income < 20,000")
income_diffs
diff_diff <- great20k_diff - less20k_diff
diff_diff

```

In this difference-in-difference design, we see that the effect of treatment on the amount in people's accounts is quite different for the lower income group compared to the higher income group. For the lower income group, we see that the effect of treatment is positive-- people in the `After Payday` group have a higher balance in their accounts than people in the `Before Payday` group (the difference is `r round(less20k_diff, 2)`. But for the higher income group, the effect of treatment appears to be negative- those in the After Payday group have `r round(abs(great20k_diff), 2)` less than those in the Before Payday group. This brings into question the idea that the study successfully manipulated economic resources through their design, especially for higher income individuals. 

**OPTIONAL FOR THE AMBITIOUS STUDENT** who compares heterogeneous effects based on mean and median estimates.

```{r}
great20k_diff2 <- median(poverty$accts_amt[poverty$income_less20k == 0 &
                         poverty$treatment == "After Payday"], na.rm = TRUE) -
  median(poverty$accts_amt[poverty$income_less20k == 0 &
                         poverty$treatment == "Before Payday"], na.rm = TRUE)
great20k_diff2

less20k_diff2 <- median(poverty$accts_amt[poverty$income_less20k == 1 &
                         poverty$treatment == "After Payday"], na.rm = TRUE) -
  median(poverty$accts_amt[poverty$income_less20k == 1 &
                         poverty$treatment == "Before Payday"], na.rm = TRUE)
great20k_diff2

diff_diff <- great20k_diff2 - less20k_diff2

```

Notice that estimates based on the median do not show differential effects of payday on amount in checking and saving accounts for groups earning more vs. less than 20,000 dollars a year. In order for the authors to convince us that they have successfully manipulated financial resources, and that their main fiinding--that poverty does not affect financial decision-making--has support, they would have to conduct all their analyses based on the median and deal with the outliers in their data.
