---
title: "International Trade Network"
output:
  pdf_document: default
  html_document: default
date: "5 August 2015"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(error=FALSE,comment='')
```

The size and structure of international trade flows varies
significantly over time. This exercise is based in part on
  Luca De Benedictis and Lucia Tajoli. (2011). 'The World Trade
  Network.' *The World Economy*, 34:8, pp.1417-1454. The trade data 
  are from Katherine Barbieri and Omar
  Keshk. (2012). *Correlates of War Project Trade Data Set*,
  Version 3.0. available at 
  [http://correlatesofwar.org](http://correlatesofwar.org).
  
The volume of goods traded between countries has grown rapidly over the
past century, as technological advances lowered the cost of shipping
and countries adopted more liberal trade policies.  At times, however,
trade flows have decreased due to disruptive events such as major wars
and the adoption of protectionist trade policies.  In this exercise,
we will explore some of these changes by examining the network of
international trade over several time periods.  The data file
`trade.csv` contains the value of exports from one country to
another in a given year.  The names and descriptions of variables in
this data set are:

--------------------------------------------------------------------------------
 Name                 Description
 -------------------- ----------------------------------------------------------
 `country1`           Country name of exporter 
 
 `country2`           Country name of importer
 
 `year`               Year
 
 `exports`            Total value of exports (in tens of millions of dollars)
--------------------------------------------------------------------------------

The data are given for years 1900, 1920, 1940, 1955, 1980, 2000, and 2009.

## Question 1

We begin by analyzing international trade as an unweighted,
  directed network.  For every year in the data set, create an
  adjacency matrix whose entry $(i,j)$ equals 1 if country $i$ exports
  to country $j$.  If this export is zero, then the entry equals 0.
  We assume that missing data, indicated by `NA`, represents
  zero trade.  Plot the 'network density', which is defined over
  time as follows,
  $$
    \text{network density}  =  \frac{\text{number of edges}}{\text{number of potential edges}}
  $$
  The `graph.density` function can compute this measure given
  an adjacency matrix.  Interpret the result.  
  
## Answer 1

```{r}
par(cex = 1.5)
trade <- read.csv("data/trade.csv", stringsAsFactors = FALSE)

library(igraph)
## change NA to 0
trade$exports[is.na(trade$exports)] <- 0
## Create a loop over three years in the dataset
years <- unique(trade$year)
densities <- rep(NA, length(years))
adj.mat <- list() # create an empty list
for(i in 1:length(years)){
    ## Subset the dataset
    trade.year <- trade[trade$year == years[i],]
    ## Create a vector of all unique countries in this year
    countries <- unique(c(trade.year$country1, 
                          trade.year$country2))
    ## Construct an unweighted, directed adjacency matrix 
    adj.year <- matrix(0, nrow = length(countries), 
                       ncol = length(countries))
    rownames(adj.year) <- colnames(adj.year) <- countries
    for (j in 1:nrow(trade.year)){
        exporter <- trade.year$country1[j]
        importer <- trade.year$country2[j]
        adj.year[exporter, importer] <- trade.year$exports[j]
    }
    ## unweighted network
    adj.year <- ifelse(adj.year > 0, 1, 0)
    ## Create network object and calculate graph density
    net <- graph.adjacency(adj.year, diag = FALSE)
    densities[i] <- graph.density(net)
    adj.mat[[i]] <- net # save network for each year
}
names(adj.mat) <- years
## Plot network density over time     
plot(years, densities, type = "b",
     main = "Density of International Trade Network")
```

The trade network has generally grown more dense from the period from
1900 to 2009.  This means that over time, countries tend to gain more
trade partners, exporting to a higher proportion of potential trade
partners.  However, there is a notable decrease in density in the year
1940, most likely reflecting protectionist policies adopted in the
inter-war period.

## Question 2

For the years 1900, 1955, and 2009, compute the measures of
  centrality based on degree, betweenness, and closeness (based on
  total degree) for each year.  For each year, list the five countries
  that have the largest values of these centrality measures.  How do
  the countries on the lists change over time?  Briefly comment on the
  results.

## Answer 2

```{r}
## Centrality measures for 1900, 1955, 2009
degree.1900 <- degree(adj.mat[["1900"]])
between.1900 <- betweenness(adj.mat[["1900"]])
close.1900 <- closeness(adj.mat[["1900"]], mode = "all")
degree.1955 <- degree(adj.mat[["1955"]])
between.1955 <- betweenness(adj.mat[["1955"]])
close.1955 <- closeness(adj.mat[["1955"]], mode = "all")
degree.2009 <- degree(adj.mat[["2009"]])
between.2009 <- betweenness(adj.mat[["2009"]])
close.2009 <- closeness(adj.mat[["2009"]], mode = "all")
## Identify 5 countries that rank highest on each
sort(degree.1900, decreasing = TRUE)[1:5]
sort(between.1900, decreasing = TRUE)[1:5]
sort(close.1900, decreasing = TRUE)[1:5]
sort(degree.1955, decreasing = TRUE)[1:5]
sort(between.1955, decreasing = TRUE)[1:5]
sort(close.1955, decreasing = TRUE)[1:5]
sort(degree.2009, decreasing = TRUE)[1:5]
sort(between.2009, decreasing = TRUE)[1:5]
sort(close.2009, decreasing = TRUE)[1:5]
```

In earlier periods, Western countries (e.g., the United Kingdom,
Germany, the United States) tended to dominate the international trade
network in terms of both closeness and betweenness centrality.  By
2009, however, the top 5 countries had grown significantly more
diverse.  For example, at that time, China and India were the two
top-ranking countries in closeness centrality, and Taiwan was the
top-ranking country in betweenness centrality.

## Question 3
  
We now analyze the international trade network as a weighted,
  directed network in which each edge has a non-negative weight
  proportional to its corresponding trade volume.  Create an adjacency
  matrix for such network data.  For the years 1900, 1955, and 2009,
  compute the centrality measures from above for the weighted trade
  network.  Instead of degree, however, compute the *graph strength*, 
  which in this case equals the sum of imports and
  exports with all adjacent nodes.  The `graph.strength`
  function can be used to compute this weighted version of degree.
  For betweenness and closeness, we use the same function as before,
  i.e., `closeness` and `betweenness`, which can handle
  weighted graphs appropriately.  Do the results differ from those of
  the unweighted network?  Examine the top five countries.  Can you
  think of another way to calculate centrality in this network that
  accounts for the value of exports from each country?  Briefly
  discuss.
  
## Answer 3

```{r}
## the following code is similar to what we had earlier
wadj.mat <- list() # create an empty list
years <- c(1900, 1955, 2009)
for(i in 1:length(years)){
    ## Subset the dataset
    trade.year <- trade[trade$year == years[i],]
    ## Create a vector of all unique countries in this year
    countries <- unique(c(trade.year$country1, 
                          trade.year$country2))
    ## Construct an unweighted, directed adjacency matrix 
    adj.year <- matrix(0, nrow = length(countries), 
                       ncol = length(countries))
    rownames(adj.year) <- colnames(adj.year) <- countries
    for (j in 1:nrow(trade.year)) {
        exporter <- trade.year$country1[j]
        importer <- trade.year$country2[j]
        adj.year[exporter, importer] <- trade.year$exports[j]
    }
    ## Create network object and calculate graph density
    wadj.mat[[i]] <- graph.adjacency(adj.year, weighted = TRUE, 
                                     diag = FALSE)
}
names(wadj.mat) <- years
## Centrality measures for 1900, 1955, 2009
strength.1900 <- graph.strength(wadj.mat[["1900"]])
between.1900 <- betweenness(wadj.mat[["1900"]])
close.1900 <- closeness(wadj.mat[["1900"]], mode = "all")
strength.1955 <- graph.strength(wadj.mat[["1955"]])
between.1955 <- betweenness(wadj.mat[["1955"]])
close.1955 <- closeness(wadj.mat[["1955"]], mode = "all")
strength.2009 <- graph.strength(wadj.mat[["2009"]])
between.2009 <- betweenness(wadj.mat[["2009"]])
close.2009 <- closeness(wadj.mat[["2009"]], mode = "all")
## Identify 5 countries that rank highest on each
sort(strength.1900, decreasing = TRUE)[1:5]
sort(between.1900, decreasing = TRUE)[1:5]
sort(close.1900, decreasing = TRUE)[1:5]
sort(strength.1955, decreasing = TRUE)[1:5]
sort(between.1955, decreasing = TRUE)[1:5]
sort(close.1955, decreasing = TRUE)[1:5]
sort(strength.2009, decreasing = TRUE)[1:5]
sort(between.2009, decreasing = TRUE)[1:5]
sort(close.2009, decreasing = TRUE)[1:5]
```

The betweenness and closeness centrality measures for the weighted
networks produce results substantially different from those of the
unweighted networks.  In particular, they assign high betweenness and
closeness centrality scores to countries that do not have large
economies, such as Qatar and Albania.  These scores suggest that while
these countries may not trade in large quantities themselves, they
link to other large trading partners that would otherwise be distantly
connected.  A weighted version of degree centrality shows that the
United States and other Western countries have played a central role
in the world trade network.

## Question 4

Apply the PageRank algorithm to the weighted trade network
  separately for each year.  For each year, identify the 5 most
  influential countries according to this algorithm.  In addition,
  examine how the ranking of PageRank values has changed over time for
  each of the following five countries -- US, United Kingdom, Russia,
  Japan, and China.  Briefly comment on the patterns you observe.
  
## Answer 4

```{r}
## PageRank measures 
pr.scores <- matrix(NA, nrow = length(years), ncol = 5)
colnames(pr.scores) <- c("United States of America", "United Kingdom", 
                         "Russia", "Japan", "China")
rownames(pr.scores) <- years
for (i in 1:length(years)) {
    pr <- page.rank(wadj.mat[[i]])$vector
    cat("year", years[i], ":\n")
    print(sort(pr, decreasing = TRUE)[1:5])
    pr.order <- names(pr)[order(pr, decreasing = TRUE)]
    pr.scores[i, ] <- match(colnames(pr.scores), pr.order) 
}
pr.scores
```

The United States has had the highest PageRank since 1940.  Prior to
then, United Kingdom was the most influential according to the
PageRank algorithm. In recent years, the rankings of Japan and then
China have risen according to PageRank values while the rankings of
countries like the United Kingdom and Russia have fallen.  Western
European countries like Germany and France remain among the most
influential countries.
