---
title: "List Experiment and Racial Prejudice"
output:
  html_document: default
  pdf_document: default
---

Despite the legal end of desegregation in the United States South,
racial tension persists. While outright discrimination against
African-Americans is illegal, some believe that many white Southerners
continue to be prejudiced against blacks at higher rates than whites
in the rest of the United States. Others have suggested that, during
the 1970s and 1980s, a 'new South' emerged, in which racial tensions
declined and the attitudes of white Southerners came to closely mirror
the rest of the country. Because of the sensitive nature of racial
prejudice, however, asking respondents directly about their feelings
towards blacks is likely to elicit untruthful answers from some
people. To get around this, researchers used a list experiment to
estimate the proportion of respondents who exhibit racial prejudice.
We saw the application of list experiment in
Section 3.1 where it was used to measure support
for combatants among Afghan citizens.  Readers are encouraged to read
that section before they begin this exercise.

In the 1991 National Race and Politics survey, researchers randomly
divided respondents into two groups.  Those assigned to the control
condition were read the following script: 

> Now, I am going to read you three things that sometimes make people angry or upset. After I read all three, just tell me HOW MANY of them upset you. I don't want to know which ones, just HOW MANY.
>
> - The federal government increasing the tax on gasoline
> - Professional athletes getting million-dollar contracts
> - Large corporations polluting the environment

Those respondents assigned to the treatment group, on the other hand,
received the same script except that the list included one additional
item that read 'a black family moving in next door.' The data set we
will be analyzing for this exercise is contained in the csv file
*listexp.csv*. The names and descriptions of the variables
in this data set are listed in the table below.

------------------------------------------------------------------------------
 Name                Description
 ------------------- ---------------------------------------------------------
 `id`                Unique id of the respondent
 
 `y`                 The number of items selected from the list
 
 `treat`             The treatment indicator (binary)
 
 `south`             An indicator for residence in a Southern state
------------------------------------------------------------------------------

## Question 1

Begin by computing the overall proportion of respondents who
  demonstrate racial prejudice. For now, remove all observations for
  which there are missing values on the outcome variable.  Compute the
  standard error and 95% confidence interval for this estimate.  Give
  a brief interpretation of the result.  
  
## Answer 1

```{r}
race <- read.csv("data/listexp.csv", header = TRUE)
race.c <- na.omit(race)

items.treat <- race.c$y[race.c$treat == 1]
items.control <- race.c$y[race.c$treat == 0]
prop.prej <- mean(items.treat) - mean(items.control)
se.prej <- sqrt(var(items.treat) / length(items.treat) + 
                  var(items.control) / length(items.control))
ci.prej <- c(prop.prej - qnorm(0.975)*se.prej, 
             prop.prej + qnorm(0.975)*se.prej)
list(prop.prej, ci.prej)
```

We find that the standard error is `r round(se.prej, digits = 2)` and the 95% confidence
interval is `r round(ci.prej, digits = 2)`. The confidence interval is telling us
that if we were to run this experiment many times, the 95% interval 
would include the true sample proportion of those who hold
racial prejudice 95% of the time. Note that this is not the same as an interval
that contains the true sample proportion with 95% probability. 
 
## Question 2

Conduct a two-sided hypothesis test where the null hypothesis is
  that the population proportion of respondents exhibiting racial
  prejudice is zero. Calculate the z-score and its associated
  two-sided $p$-value.  Finally, conduct the hypothesis test using
  0.05 as the statistical significance threshold. What assumptions are 
  required in order for this estimate to be valid?
  
## Answer 2

```{r}
## Remember, the real numerator is prop.prej - 0 because of null hypothesis
z.prej <- prop.prej/se.prej 
## Set lower.tail to get the correct portion of the distribution
p.val <- 2*pnorm(z.prej, lower.tail = F) 
```

We obtain a z-score of `r round(z.prej, digits = 2)` and an associated $p$-value of 
`r round(p.val, digits = 2)`. At the 0.05 level of significance, we fail to 
reject the null hypothesis of no treatment effect. This means that the difference 
in the mean number of items selected in the treatment condition is not statistically 
significant. Substantively, this means that we cannot reject the claim that there is 
no racial prejudice in the survey sample. In obtaining the p-value, we are assuming that
our test statistic (e.g. the difference in means between treated and control) is normally 
distributed. If the $t$-statistic was used, we would be assuming that the difference
is distributed according to the Student's $t$ distribution with $n-1$ degrees of freedom.

## Question 3

Conduct the same hypothesis test as in the previous question 
  separately for Southern and non-Southern respondents. Test the null
  hypothesis of zero difference between the proportion of respondents
  exhibiting racial prejudice between the Southern and non-Southern
  respondents. Report the $p$-values under the alternative hypothesis
  that the population proportion of respondents exhibiting racial
  prejudice is greater in the Southern sample than in the non-Southern 
  sample.  Interpret the results of this hypothesis test.
  
## Answer 3

```{r}
## South
items.treat.south <- race.c$y[race.c$treat == 1 & race.c$south == 1]
items.control.south <- race.c$y[race.c$treat == 0 & race.c$south == 1]
prop.prej.south <- mean(items.treat.south) - mean(items.control.south)
se.prej.south <- sqrt(var(items.treat.south) / length(items.treat.south) + 
                  var(items.control.south) / length(items.control.south))
z.prej.south <- prop.prej.south / se.prej.south
pnorm(z.prej.south, lower.tail = F)

## Non-South

items.treat.nosouth <- race.c$y[race.c$treat == 1 & race.c$south == 0]
items.control.nosouth <- race.c$y[race.c$treat == 0 & race.c$south == 0]
prop.prej.nosouth <- mean(items.treat.nosouth) - mean(items.control.nosouth)
se.prej.nosouth <- sqrt(var(items.treat.nosouth) / length(items.treat.nosouth) + 
                  var(items.control.nosouth) / length(items.control.nosouth))
z.prej.nosouth <- prop.prej.nosouth / se.prej.nosouth
pnorm(z.prej.nosouth, lower.tail = F)

## Difference
treat.diff <- prop.prej.south - prop.prej.nosouth
se.diff <- sqrt(se.prej.south^2 + se.prej.nosouth^2)
p.diff <- pnorm(treat.diff / se.diff, lower.tail = F)

```

The difference in proportions for the South and non-South samples
is `r round(treat.diff, digits = 2)`, indicating that a larger share of respondents
in the South exhibit racial prejudice (in the sample). The $p$-value 
of the z-score of the difference between the South and non-South
respondents is `r round(p.diff, digits = 3)`, that this difference is statistically
significant at the 0.05 level. The interpretation of this result
is that if we conducted the experiment many times, the  probability
that we would observe a difference as large as we do is `r round(p.diff*100, digits = 3)`. 
We can thus reject the null hypothesis that the incidence of racial
prejudice is the same in both samples. 

## Question 4

Construct the 95 percent confidence interval for the
  difference in the population proportion of those who are prejudiced
  between the South and non-Southern whites.  Interpret the result.
  
## Answer 4

```{r}
ci.south <- c(prop.prej.south + qnorm(0.975)*se.prej.south, 
              prop.prej.south - qnorm(0.975)*se.prej.south)
ci.nosouth <- c(prop.prej.nosouth + qnorm(0.975)*se.prej.nosouth, 
              prop.prej.nosouth - qnorm(0.975)*se.prej.nosouth)
ci.diff <- c(treat.diff + qnorm(0.975)*se.diff, 
             treat.diff - qnorm(0.975)*se.diff)
```

We see that the confidence interval is quite large, ranging from a difference
of slightly less than 1% to a difference of nearly 50%. As such, while
the difference we detect is statistically significant, it is also relatively 
imprecise. The true difference could be of a magnitude that is either substantively very large (`r ci.diff[1]`) or very small (`r ci.diff[2]`). 

## Question 5

A critical assumption of the list experiment is that the
  inclusion of the sensitive item does not alter the respondents'
  willingness to give a truthful answer to the number of items that
  upset them.  Test this assumption by examining the rates of
  non-response. Compare the Southern and non-Southern subsets.  Are
  there differences between the Southern and non-Southern respondents?
  What does this tell us about the validity of the list experiment and
  racial attitudes between the regions?
  
## Answer 5

```{r}
race$missing <- ifelse(is.na(race$y) == T, 1, 0)

## South
miss.treat.south <- race$miss[race$treat == 1 & race$south == 1]
miss.control.south <- race$miss[race$treat == 0 & race$south == 1]
prop.miss.south <- mean(miss.treat.south) - mean(miss.control.south)
se.miss.south <- sqrt(var(miss.treat.south) / length(miss.treat.south) + 
                  var(miss.control.south) / length(miss.control.south))
z.miss.south <- prop.miss.south / se.miss.south
pnorm(z.miss.south, lower.tail = F)

## Non-South

miss.treat.nosouth <- race$miss[race$treat == 1 & race$south == 0]
miss.control.nosouth <- race$miss[race$treat == 0 & race$south == 0]
prop.miss.nosouth <- mean(miss.treat.nosouth) - mean(miss.control.nosouth)
se.miss.nosouth <- sqrt(var(miss.treat.nosouth) / length(miss.treat.nosouth) + 
                  var(miss.control.nosouth) / length(miss.control.nosouth))
z.miss.nosouth <- prop.miss.nosouth / se.miss.nosouth
pnorm(z.miss.nosouth, lower.tail = F)

## Difference
treat.diff <- prop.miss.south - prop.miss.nosouth
se.diff <- sqrt(se.miss.south^2 + se.miss.nosouth^2)
p.val <- pnorm(treat.diff / se.diff, lower.tail = F)
```

There do not appear to be any statistically significant differences in 
the non-response rates across the two subsets. The estimated difference of
`r round(treat.diff, digits=4)` is 
substantively very small and the large $p$-value of 
`r round(p.val, digits=4)` means that we cannot reject the null 
hypothesis of no 
difference in non-response rates across the two samples. This strengthens
the validity of the list experiment because it suggests that
people did not react differently to the inclusion of the sensitive item
depending on what region they were from. 

## Question 6

Now we conduct randomization inference separately for the South
  and non-South samples. Using the difference-in-means estimator as a
  test statistic, simulate the randomization distribution for 10,000
  possible treatment assignments under the null hypothesis that the
  population proportion of those who are racially prejudiced is zero.
  For both the South and non-South samples, create a histogram of this
  randomization distribution and include as a vertical line the actual
  estimated proportion. Then, use these distributions to approximate
  the one-sided $p$-values and conduct the hypothesis test using 0.05 as the
  threshold of statistical significance.  Provide a brief interpretation 
  of the result.
  
## Answer 6

```{r}
## Store the estimated ATE for both South and non-South region:
ate <- c(prop.prej.nosouth, prop.prej.south)
## Subset data
race.south <- race[race$south == 1, ]
race.nosouth <- race[race$south == 0, ]
## Specify number of simulations
n.sims <- 10000
## Create randomization distributions
rand.south <- rep(NA, n.sims)
rand.nosouth <- rep(NA, n.sims)

## Begin loop
for(i in 1:n.sims){
## Build random treatment assignment vector
  treat <- sample(race.c$treat, nrow(race.c), replace = FALSE) 
  race.c$treat <- treat
## Compute means and store
  rand.south[i] <- mean(race.c$y[race.c$treat == 1 & race.c$south == 1], na.rm = T) - 
                    mean(race.south$y[race.c$treat == 0 & race.c$south == 1], na.rm = T)
  rand.nosouth[i] <- mean(race.c$y[race.c$treat == 1 & race.c$south == 0], na.rm = T) - 
                    mean(race.c$y[race.c$treat == 0 & race.c$south == 0], na.rm = T)
}

## Visualize in histograms
par(mfrow = c(1,2))
hist(rand.nosouth, freq = FALSE, breaks = 20,
     xlab = "Estimated Proportion Exhibiting Prejudice", 
     main = "Non-Southern Respondents")
abline(v = ate[1], lty = 2, col = "red")
hist(rand.south, freq = FALSE, breaks = 20, 
     xlab = "Estimated Proportion Exhibiting Prejudice", 
     main = "Southern Respondents")
abline(v = ate[2], lty = 2, col = "red")

## Get exact one-sided p-value: 
exact.nosouth <- length(rand.nosouth[rand.nosouth >= ate[1]])/length(rand.nosouth)
exact.south <- length(rand.south[rand.south >= ate[2]])/length(rand.south)
```

The results of the randomization inference analysis is very similar to the analysis above
that relies on the normality assumption to obtain the p-value. However, the interpretation
of the $p$-value is different under randomization distribution, as is the null hypothesis. 
Here, we are testing the sharp null hypothesis of no treatment effect for all units, which is
different than in our previous analysis, where we tested the null hypothesis of no difference, 
on average, between treated and untreated units. The exact $p$-values of `r format(exact.nosouth, scientific=FALSE)`
and `r format(exact.south, scientific=FALSE)` for the non-South and South sample, respectively, represent the probability
that we would have observed the treatment effect that we did purely because of the chosen
treatment assignment vector under the sharp null hypothesis. The small $p$-value for the South
sample means that it is extremely unlikely we would have observed as large an effect as we did
if the sharp null hypothesis were true. We can thus reject the sharp null for the Southern sample
but not for the non-Southern sample. 
